### 2024-01-22 outputs/paper.qmd
Initial paper

---
title: "This is the title"
abstract: "This is the abstract"
thanks: "Code and data supporting this analysis is available at: https://github.com/prajogt/ttc_analysis.git"
author: "Timothius Prajogi"
date: today
date-format: long
format: pdf
editor: visual
toc: true
number-sections: true
bibliography: refrences.bib
---

```{r}
#| echo: false
#| message: false
#| warning: false

#### Workspace setup ####
library(janitor)
library(knitr)
library(opendatatoronto)
library(tidyverse)

```

## Introduction

## Data

Cite the packages used here.

Datasets used: subway data [@citesubwaydata], bus data [@citebusdata]

```{r}
#| echo: false
#| message: false
#| warning: false

bus_delay_data <- 
  read_csv(
     file = "data/cleaned_bus_delay_statistics.csv",
     show_col_types = FALSE
  )

subway_delay_data <-
  read_csv(
     file = "data/cleaned_subway_delay_statistics.csv",
     show_col_types = FALSE
  )
```

## Results

## Discussion

## Conclusion

## Reference

### 2024-01-22 scripts/00-simulate_data.R
Update probabilities based on days

@@ -11,7 +11,9 @@
 library(tidyverse)
 
 
-# Simulated situation attributes
+## Simulated situation attributes ##
+
+# The possible incident types causing the delay
 sim_incident_types <- 
   c(
     "Equipment/Mechanical", 
@@ -29,6 +31,7 @@ sim_incident_probabilities <-
     0.2
   )
 
+# The days of the week the buses and subways operate
 sim_operating_days <-
   c(
     "Sunday",
@@ -43,18 +46,23 @@ sim_operating_days <-
 # Assumption that delays change depending on the day
 sim_probability_of_subway_delay <-
   c(
-    0.44, # Sort of high on Sundays
-    0.33, # Slightly lower as people go back to school/work
-    0.21, # Low on Tuesday
-    0.30, # Low on Wednesday
-    0.27, # Low on Thursday
-    0.51, # High on Friday
-    0.54  # High on Saturday
+    0.18, # Sort of high on Sundays
+    0.14, # Slightly lower as people go back to school/work
+    0.10, # Low on Tuesday
+    0.10, # Low on Wednesday
+    0.11, # Low on Thursday
+    0.17, # High on Friday
+    0.20  # High on Saturday
   )
+
+# Sample 1 time from a possion dist with an average
+# of 20,000 delays in 1 day
+num_subway_delays <- rpois(1, 20,000)
 
 # Assumption that there are more bus delays (strong base since bus delay 
 # dataset is twice the size) and subway delays and bus delays are correlated
-sim_probability_of_bus_delay <- 1.5 * sim_probability_of_subway_delay
+num_bus_delays <- 1.5 * num_subway_delays
+ 
 
 
 # Simulate delays happening

### 2024-01-22 scripts/00-simulate_data.R
Complete simulations for the data
TODO plotting the simulated data

@@ -9,6 +9,21 @@
 
 #### Workspace setup ####
 library(tidyverse)
+#### Preamble ####
+# Purpose: Simulate delay data
+# Author: Timothius Prajogi
+# Date: 23 January 2024
+# Contact: tim.prajogi@mail.utoronto.ca
+# License: MIT
+# Prerequisites: None
+
+
+#### Workspace setup ####
+library(tidyverse)
+set.seed(302)
+
+# Clean all objects from the workspace
+rm()
 
 
 ## Simulated situation attributes ##
@@ -57,7 +72,60 @@ sim_probability_of_subway_delay <-
 
 # Sample 1 time from a possion dist with an average
 # of 20,000 delays in 1 day
-num_subway_delays <- rpois(1, 20,000)
+num_subway_delays <- rpois(1, 20000)
+
+# Assumption that there are more bus delays (strong base since bus delay 
+# dataset is twice the size) and subway delays and bus delays are correlated
+num_bus_delays <- 1.5 * num_subway_delays
+
+
+## Simulated situation attributes ##
+
+# The possible incident types causing the delay
+sim_incident_types <- 
+  c(
+    "Equipment/Mechanical", 
+    "Miscellaneous", 
+    "Security/Safety", 
+    "Operator"
+  )
+
+# Expected distribution of incident types
+sim_incident_probabilities <-
+  c(
+    0.3,
+    0.2,
+    0.4,
+    0.2
+  )
+
+# The days of the week the buses and subways operate
+sim_operating_days <-
+  c(
+    "Sunday",
+    "Monday",
+    "Tuesday",
+    "Wednesday",
+    "Thursday",
+    "Friday",
+    "Saturday"
+  )
+
+# Assumption that delays change depending on the day
+sim_probability_of_subway_delay <-
+  c(
+    0.18, # Sort of high on Sundays
+    0.14, # Slightly lower as people go back to school/work
+    0.10, # Low on Tuesday
+    0.10, # Low on Wednesday
+    0.11, # Low on Thursday
+    0.17, # High on Friday
+    0.20  # High on Saturday
+  )
+
+# Sample 1 time from a possion dist with an average
+# of 20,000 delays in 1 day
+num_subway_delays <- rpois(1, 20000)
 
 # Assumption that there are more bus delays (strong base since bus delay 
 # dataset is twice the size) and subway delays and bus delays are correlated
[…]

### 2024-01-23 scripts/04-summarise_data.R
Add summaries

@@ -1,0 +1,108 @@
+#### Preamble ####
+# Purpose: Group delay data into similar sets of incidents
+# Author: Timothius Prajogi
+# Date: 23 January 2024
+# Contact: tim.prajogi@mail.utoronto.ca
+# License: MIT
+# Prerequisites: 01-download_data, 02-clean_data
+
+#### Workspace setup ####
+library(tidyverse)
+library(lubridate)
+
+# Read in bus data
+bus_delay_data <- 
+  read_csv(
+    file = "outputs/data/cleaned_bus_delay_statistics.csv",
+    show_col_types = FALSE
+  )
+
+# Enforce ordering of days
+bus_delay_data$day <- factor(bus_delay_data$day, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
+
+# Summarize into average number of delays per day of the week (Mon, Tues, ...)
+avg_num_bus_delays_by_day <-
+  bus_delay_data |> 
+  group_by(date, day) |> 
+  summarise(n = n(), .groups = "keep") |>
+  group_by(day) |>
+  summarise(mean_num_delays = mean(n))
+
+# Summarize into total delay time and number of delays per date (Jan 1st, 2nd, ...)
+total_bus_delay_time_by_date <-
+  bus_delay_data |> 
+  group_by(date, day) |> 
+  summarise(
+    total_delay_time = sum(min_delay), 
+    n = n(), 
+    mean_delay_time = (total_delay_time / n),
+    .groups = "keep")
+
+# Summarize into average delay time per day of the week
+avg_bus_delay_time_by_day <-
+  total_bus_delay_time_by_date |>
+  group_by(day) |>
+  summarise(mean = mean(total_delay_time))
+
+# Summarize into the total number of delays per incident type (e.g. Equipment/Mechanical)
+total_num_bus_delays_by_incident <-
+  bus_delay_data |>
+  group_by(incident) |>
+  summarise(n = n())
+
+# Read in subway data
+subway_delay_data <- 
+  read_csv(
+    file = "outputs/data/cleaned_subway_delay_statistics.csv",
+    show_col_types = FALSE
+  )
+
+# Enforce ordering of days
+subway_delay_data$day <- factor(subway_delay_data$day, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
+
[…]

### 2024-01-23 scripts/00-simulate_data.R
Clean up workspace after the fact

@@ -22,10 +22,6 @@ library(tidyverse)
 library(tidyverse)
 set.seed(302)
 
-# Clean all objects from the workspace
-rm()
-
-
 ## Simulated situation attributes ##
 
 # The possible incident types causing the delay
@@ -216,3 +212,5 @@ sim_data_bus$delay_time |>
 sim_data_bus$delay_time |>
   max() <= 60 * 5
 
+# Clean up workspace
+rm(list = ls())

### 2024-01-23 scripts/01-download_data.R

@@ -11,9 +11,6 @@
 library(opendatatoronto)
 library(tidyverse)
 
-# Clean all objects from the workspace
-rm()
-
 #### Download and save TTC data ####
 
 ## Subway delay data
@@ -48,5 +45,7 @@ raw_bus_data <-
 
 write_csv(raw_bus_data, 'inputs/data/raw_bus_delay_statistics.csv')
 
+# Clean up workspace
+rm(list = ls())
 
 

### 2024-01-23 scripts/02-clean_data.R

@@ -10,9 +10,6 @@
 library(tidyverse)
 library(janitor)
 
-# Clean all objects from the workspace
-rm()
-
 #### Read and clean raw data ####
 
 ## Read in Bus data
@@ -122,4 +119,7 @@ cleaned_subway_codes <-
 
 # Save cleaned code mappings
 write_csv(cleaned_subway_codes, 'inputs/data/subway_delay_codes.csv')
+
+# Clean up workspace
+rm(list = ls())
 

### 2024-01-23 scripts/03-parse_codes.R

@@ -10,9 +10,6 @@
 library(tidyverse)
 library(janitor)
 
-# Clean all objects from the workspace
-rm()
-
 ### Bus data ###
 
 # Load in non-grouped data
@@ -60,7 +57,7 @@ subway_codes <-
 
 # Section codes into incident groups
 cleaned_subway_codes <- 
-  cleaned_subway_codes |>
+  subway_codes |>
   mutate(
     incident = case_when(
       startsWith(code, "E") ~ "Equipment/Mechanical",
@@ -109,3 +106,44 @@ cleaned_subway_delay_data <-
 write_csv(cleaned_subway_delay_data, 'outputs/data/cleaned_subway_delay_statistics.csv')
 
 
+#### Data Validation ####
+
+# Test that the incident types are limited to the set
+cleaned_subway_delay_data$incident |> 
+  unique() |>
+  setequal(c("Equipment/Mechanical", "Miscellaneous", "Operator", "Security/Safety"))
+
+cleaned_bus_delay_data$incident |> 
+  unique() |>
+  setequal(c("Equipment/Mechanical", "Miscellaneous", "Operator", "Security/Safety"))
+
+
+# Test that there are only 7 unique days
+cleaned_subway_delay_data$day |>
+  unique() |>
+  length() == 7
+
+cleaned_bus_delay_data$day |>
+  unique() |>
+  length() == 7
+
+# Test that delay times are non-negative
+cleaned_subway_delay_data$min_delay |>
+  min() >= 0
+
+cleaned_bus_delay_data$min_delay |>
+  min() >= 0
+
+# Verify datatypes
+class(cleaned_subway_delay_data$day) == "character"
+class(cleaned_subway_delay_data$incident) == "character"
+class(cleaned_subway_delay_data$min_delay) == "numeric"
+class(cleaned_subway_delay_data$min_gap) == "numeric"
+
+class(cleaned_bus_delay_data$day) == "character"
+class(cleaned_bus_delay_data$incident) == "character"
+class(cleaned_bus_delay_data$min_delay) == "numeric"
+class(cleaned_bus_delay_data$min_gap) == "numeric"
+
+# Clean up workspace
+rm(list = ls())

### 2024-01-23 outputs/paper.qmd
Add intro

@@
 
+The main source public transportation of public transportation in the city of 
+Toronto is the TTC (Toronto Transit Commission). The most common methods of
+this public transportation is subways and buses. For people who frequently
+use these methods, delays are common and expected [@CityNewsBGrade]. 
+So much so that leaving early is leaving on time.
+
+With the delays getting worse and service becoming more inconsistent 
+[@TorontoStarTTCDelays], it is important to analyze potential causes and 
+therefore enable the TTC to explore potential solutions. One of such factors
+is the tight coupling of the subway and bus systems. When issues occur with the
+subway, the overflow is sent to buses, either from detoured commuter traffic,
+or resource reallocation in the form of converting regular route buses to 
+shuttle buses [@TorontoObserverShuttleBuses].
+
+Within this paper we will analyze the correlations between the delays occurring
+on the subway and the delays that occur for buses. In particular, the 
+correlation between mechanical or security issues on the subway, which tend to 
+shut down a portion of the track, and with diversion and other external 
+issues (here after labeled "Miscellaneous") that occur for buses. A logical
+correlation was made above, and a correlation shown in data would go to further
+a cause and effect relationship.
+
+Later we will see the general affects of the commuter load on the transit system
+and how the system can be better optimized and fail-safed to prevent such frequent
+delays. The paper will show the trends of the amount of delays for both
+systems as a function of day and time. It was found that work days put additional
+load on both systems, the total amount of delays during weekdays being 
+significantly higher than on other days.
+
+The datasets used for this analysis are described in the Data section, as well
+as the process that was done to clean the data into comparable datasets.
+
 ## Data
@@
 
+# Read in cleaned data
 bus_delay_data <- 

### 2024-01-23 scripts/00-simulate_data.R
Update parsing

@@ -9,6 +9,8 @@
 
 #### Workspace setup ####
 library(tidyverse)
+
+
 #### Preamble ####
 # Purpose: Simulate delay data
 # Author: Timothius Prajogi
@@ -138,6 +140,11 @@ sim_data_subway <-
       prob = sim_probability_of_subway_delay, 
       replace = TRUE
     ),
+    "time" = rnorm(
+      num_subway_delays,
+      mean = 17, # Assuming a peak during rush hour
+      sd = 6 
+    )%% 24, # Staying within military time
     "incident" = sample(
       sim_incident_types, 
       num_subway_delays, 
@@ -158,6 +165,11 @@ sim_data_bus <-
       prob = sim_probability_of_subway_delay, 
       replace = TRUE
     ),
+    "time" = rnorm(
+      num_bus_delays,
+      mean = 17, # Assuming a peak during rush hour
+      sd = 6 
+    ) %% 24, # Staying within military time
     "incident" = sample(
       sim_incident_types, 
       num_bus_delays, 
@@ -174,9 +186,17 @@ sim_data_bus <-
 sim_data_subway
 sim_data_bus
 
-# Visualize the data
+# Visualize a sample for the data
 
-### TODO ggplot
+# Allow for display of plots side by side
+par(mfrow = c(1, 3))
+
+sim_data_subway |>
+  ggplot(aes(x=time)) + 
+  geom_histogram(binwidth = 0.2, fill="blue", alpha=0.75) +
+  ggtitle("A simulation of the amount of delays at a given time") + 
+  xlab("Time (in military time hrs)") +
+  ylab("Number of delays")
 
 
 # Testing
@@ -213,4 +233,13 @@ sim_data_bus$delay_time |>
   max() <= 60 * 5
 
 # Clean up workspace
-rm(list = ls())
+rm(list = c(
+  "num_bus_delays",
+  "num_subway_delays",
+  "sim_data_bus",
+  "sim_data_subway",
+  "sim_incident_probabilities", 
+  "sim_incident_types", 
+  "sim_operating_days",
+  "sim_probability_of_subway_delay"
+))

### 2024-01-23 scripts/01-download_data.R

@@ -46,6 +46,10 @@ raw_bus_data <-
 write_csv(raw_bus_data, 'inputs/data/raw_bus_delay_statistics.csv')
 
 # Clean up workspace
-rm(list = ls())
+rm(list = c(
+  "raw_bus_data",
+  "raw_subway_data",
+  "subway_delay_codes"
+))
 
 

### 2024-01-23 scripts/02-clean_data.R

@@ -35,6 +35,12 @@ cleaned_bus_data <-
     min_gap
   )
 
+# Filter out situations in which the delay is 0, meaning the incident that
+# occurred didn't affect service
+cleaned_bus_data <-
+  cleaned_bus_data |>
+  filter(min_gap > 0)
+
 # Save cleaned bus data
 write_csv(cleaned_bus_data, 'inputs/data/bus_delay_statistics.csv')
 
@@ -60,11 +66,27 @@ cleaned_subway_data <-
     day,
     code,
     min_delay,
-    min_gap
+    min_gap,
+    line # additional for subway (no similar grouping for buses)
   )
 
 # Save cleaned subway data
 write_csv(cleaned_subway_data, 'inputs/data/subway_delay_statistics.csv')
+
+# The reason 2 different .csv files are stored is that the vast majority
+# of rows had a delay of 0 minutes. It seems as if a delay of 0 minutes
+# means a delay of < 1 minute which is reasonable for a subway system.
+# However, since this is an assumption, I also kept the filtered 
+# dataset for reference.
+
+# Filter out situations in which the delay is 0, meaning the incident that
+# occurred didn't affect service
+cleaned_subway_data <-
+  cleaned_subway_data |>
+  filter(min_delay > 0)
+
+# Save cleaned and filtered subway data
+write_csv(cleaned_subway_data, 'inputs/data/filtered_subway_delay_statistics.csv')
 
 # Read in subway codes
 raw_subway_codes <-
@@ -120,6 +142,44 @@ cleaned_subway_codes <-
 # Save cleaned code mappings
 write_csv(cleaned_subway_codes, 'inputs/data/subway_delay_codes.csv')
 
+#### Data Validation ####
+
+# Test that there are only 7 unique days
+cleaned_subway_data$day |>
+  unique() |>
+  length() == 7
+
+cleaned_bus_data$day |>
+  unique() |>
+  length() == 7
+
+# Test that delay times are positive
+cleaned_subway_data$min_delay |>
+  min() >= 0
+
+cleaned_bus_data$min_delay |>
+  min() >= 0
+
+# Verify datatypes
+class(cleaned_subway_data$day) == "character"
+class(cleaned_subway_data$time) == c("hms", "difftime")
+class(cleaned_subway_data$min_delay) == "numeric"
[…]

### 2024-01-23 scripts/03-parse_codes.R

@@ -90,7 +90,34 @@ cleaned_subway_delay_data <-
     day,
     incident,
     min_delay,
-    min_gap
+    min_gap,
+    line
+  )
+
+# Clean up line data
+cleaned_subway_delay_data <-
+  cleaned_subway_delay_data |>
+  mutate(
+    line = case_match(line,
+      "YU" ~ "Yonge-University",
+      "YUS" ~ "Yonge-University",
+      
+      "BD" ~ "Bloor-Danforth",
+      "BD LINE 2" ~ "Bloor-Danforth",
+      
+      "SRT" ~ "Scarborough-RT",
+      
+      "SHP" ~ "Sheppard",
+      
+      "YU / BD" ~ "Yonge-University/Bloor-Danforth",
+      "BD/YU"  ~ "Yonge-University/Bloor-Danforth",
+      "YU/BD" ~ "Yonge-University/Bloor-Danforth",
+      "YUS/BD" ~ "Yonge-University/Bloor-Danforth",
+      "YU & BD" ~ "Yonge-University/Bloor-Danforth",
+      "BLOOR DANFORTH & YONGE" ~ "Yonge-University/Bloor-Danforth",
+      .default = "Other" 
+      
+    )
   )
   
 
@@ -127,7 +154,7 @@ cleaned_bus_delay_data$day |>
   unique() |>
   length() == 7
 
-# Test that delay times are non-negative
+# Test that delay times are positive
 cleaned_subway_delay_data$min_delay |>
   min() >= 0
 
@@ -136,14 +163,23 @@ cleaned_bus_delay_data$min_delay |>
 
 # Verify datatypes
 class(cleaned_subway_delay_data$day) == "character"
+class(cleaned_subway_delay_data$time) == c("hms", "difftime")
 class(cleaned_subway_delay_data$incident) == "character"
 class(cleaned_subway_delay_data$min_delay) == "numeric"
 class(cleaned_subway_delay_data$min_gap) == "numeric"
 
 class(cleaned_bus_delay_data$day) == "character"
+class(cleaned_bus_delay_data$time) == c("hms", "difftime")
 class(cleaned_bus_delay_data$incident) == "character"
 class(cleaned_bus_delay_data$min_delay) == "numeric"
 class(cleaned_bus_delay_data$min_gap) == "numeric"
 
 # Clean up workspace
-rm(list = ls())
+rm(list = c(
+  "bus_delay_data", 
[…]

### 2024-01-23 scripts/04-summarise_data.R

@@ -90,6 +90,12 @@ total_num_subway_delays_by_incident <-
   group_by(incident) |>
   summarise(n = n())
 
+# Summarize into the total number of delays per line (e.g. BY (Bloor-Yonge))
+total_num_subway_delays_by_line <-
+  subway_delay_data |>
+  group_by(line) |>
+  summarise(n = n())
+
 # Save all the data
 write_csv(avg_bus_delay_time_by_day, "outputs/data/summaries/avg_bus_delay_time_by_day.csv")
 write_csv(avg_subway_delay_time_by_day, "outputs/data/summaries/avg_subway_delay_time_by_day.csv")
@@ -103,6 +109,20 @@ write_csv(total_num_subway_delays_by_incident, "outputs/data/summaries/total_num
 write_csv(total_bus_delay_time_by_date, "outputs/data/summaries/total_bus_delay_time_by_date.csv")
 write_csv(total_subway_delay_time_by_date, "outputs/data/summaries/total_subway_delay_time_by_date.csv")
 
+write_csv(total_num_subway_delays_by_line, "outputs/data/summaries/total_num_subway_delays_by_line.csv")
+
 # Clean up workspace
-rm(list = ls())
+rm(list = c(
+  "avg_bus_delay_time_by_day",
+  "avg_num_bus_delays_by_day",
+  "avg_num_subway_delays_by_day",
+  "avg_subway_delay_time_by_day",
+  "bus_delay_data",
+  "subway_delay_data",
+  "total_bus_delay_time_by_date",
+  "total_num_bus_delays_by_incident",
+  "total_num_subway_delays_by_incident",
+  "total_num_subway_delays_by_line",
+  "total_subway_delay_time_by_date"  
+))
 

### 2024-01-23 outputs/paper.qmd
Update paper

@@
 ---
-title: "This is the title"
-abstract: "This is the abstract"
+title: "Analyzing TTC Delay Data to Improve Future Efficiency"
+abstract: "TTC delays have been increasing over time, and especially now in a period of overhaul in many parts of the system, it becomes important to analyze the effects and relations subways have on bus delays and commuter loads.
+Using data published by the TTC and accessed through `opendatatoronto` we find that the greatest causes of delays are date and time, factors of which are out of anyones control, but must be accounted for to improve future operational efficieny.
+Further analysis must be made about the correlations with subways and buses to make a stronger argument."
 thanks: "Code and data supporting this analysis is available at: https://github.com/prajogt/ttc_analysis.git"
@@
 
-The main source public transportation of public transportation in the city of 
-Toronto is the TTC (Toronto Transit Commission). The most common methods of
-this public transportation is subways and buses. For people who frequently
-use these methods, delays are common and expected [@CityNewsBGrade]. 
+The main source public transportation of public transportation in the city of Toronto is the TTC (Toronto Transit Commission).
+The most common methods of this public transportation is subways and buses.
+For people who frequently use these methods, delays are common and expected [@CityNewsBGrade].
 So much so that leaving early is leaving on time.
 
-With the delays getting worse and service becoming more inconsistent 
-[@TorontoStarTTCDelays], it is important to analyze potential causes and 
-therefore enable the TTC to explore potential solutions. One of such factors
-is the tight coupling of the subway and bus systems. When issues occur with the
-subway, the overflow is sent to buses, either from detoured commuter traffic,
-or resource reallocation in the form of converting regular route buses to 
-shuttle buses [@TorontoObserverShuttleBuses].
+With the delays getting worse and service becoming more inconsistent [@TorontoStarTTCDelays], it is important to analyze potential causes and therefore enable the TTC to explore potential solutions.
+One of such factors is the tight coupling of the subway and bus systems.
+When issues occur with the subway, the overflow is sent to buses, either from detoured commuter traffic, or resource reallocation in the form of converting regular route buses to shuttle buses [@TorontoObserverShuttleBuses].
 
-Within this paper we will analyze the correlations between the delays occurring
-on the subway and the delays that occur for buses. In particular, the 
-correlation between mechanical or security issues on the subway, which tend to 
-shut down a portion of the track, and with diversion and other external 
-issues (here after labeled "Miscellaneous") that occur for buses. A logical
-correlation was made above, and a correlation shown in data would go to further
-a cause and effect relationship.
+Within this paper we will analyze the correlations between the delays occurring on the subway and the delays that occur for buses.
+In particular, the correlation between mechanical or security issues on the subway, which tend to shut down a portion of the track, and with diversion and other external issues (here after labeled "Miscellaneous") that occur for buses.
+A logical correlation was made above, and a correlation shown in data would go to further a cause and effect relationship.
 
-Later we will see the general affects of the commuter load on the transit system
-and how the system can be better optimized and fail-safed to prevent such frequent
-delays. The paper will show the trends of the amount of delays for both
-systems as a function of day and time. It was found that work days put additional
-load on both systems, the total amount of delays during weekdays being 
-significantly higher than on other days.
+Later we will see the general affects of the commuter load on the transit system and how the system can be better optimized and fail-safed to prevent such frequent delays.
+The paper will show the trends of the amount of delays for both systems as a function of day and time.
+It was found that work days put additional load on both systems, the total amount of delays during weekdays being significantly higher than on other days.
 
-The datasets used for this analysis are described in the Data section, as well
-as the process that was done to clean the data into comparable datasets.
+The datasets used for this analysis are described in the Data section, as well as the process that was done to clean the data into comparable datasets.
 
@@
 
-Cite the packages used here.
+The datasets used for this report are the latest delay statistics from the 2023 year, `TTC SUbway Delay Data` [@citesubwaydata] and `TTC Bus Delay Data` [@citebusdata], both of which were published by the TTC.
 
-Datasets used: subway data [@citesubwaydata], bus data [@citebusdata]
+Data was downloaded, cleaned, parsed, analyzed, and visualized using R [@citeR], a statistical programming language, with package support from `tidyverse` [@citetidyverse], a collection of libraries which included the following packages that were utilized: \* `ggplot2` [@citeggplot] \* `dplyr` [@citedplyr] \* `readr` [@citereadr] \* `tibble` [@citetibble]
+
+For additional assistance with cleaning, the `janitor` [@citejanitor] package was used. For additional assistance with report generation the `knitr` [@citeknitr] package was used.
 
@@
 
+### Overview of Datasets
+
+```{r}
+#| echo: false
+#| message: false
+#| warning: false
+#| label: tbl-bus_delay_data
+#| tbl-cap: Sample of Bus Delay Statistics
+
+# Display a sample of the bus dataset in a table
+bus_delay_data |>
+  head(5) |>
+  kable(
+    col.names = c("Date", "Time", "Day", "Incident", "Min Delay", "Min Gap"),
+    booktabs = TRUE
+  )
+```
+
+```{r}
+#| echo: false
+#| message: false
+#| warning: false
+#| label: tbl-subway_delay_data
+#| tbl-cap: Sample of Subway Delay Statistics
+
+# Display a sample of the subway dataset in a table
+subway_delay_data |>
+  head(5) |>
+  kable(
+    col.names = c("Date", "Time", "Day", "Incident", "Min Delay", "Min Gap", "Line"),
+    booktabs = TRUE
+  )
+
+```
+
+Both @tbl-bus_delay_data and @tbl-subway_delay_data contain columns "Date" and "Time" in which the delay had occurred, and an "Incident" in which the type of incident that had occurred out of these 4 general types: "Operator" (for driver/operator related issues), "Security/Safety" (for fire, police, trespassing, etc.), "Equipment/Mechanical" (for equipment, track, vehicle, etc., failures), and "Miscellaneous" (for diversions and other external issues).
+
+For @tbl-bus_delay_data and @tbl-subway_delay_data the "Min Delay" and "Min Gap" are in minutes.
+Where "Min Delay" represents the delay in the schedule for the next bus and "Min Gap" is the scheduled time from the bus ahead of the following bus.
+
+For @tbl-subway_delay_data the additional parameter "Line" represents the subway line in which the delay took place in. For those not familiar
+
+In total @tbl-bus_delay_data contains 51,406 entries, whilst @tbl-subway_delay_data contains 22,465 entries.
+
+Both datasets went through basic cleaning for readability and irrelevant or unusable columns were removed and not considered in this report.
+
+Originally, the subway data reported the incident causing the delay as a unique code rather than an understandable description.
+This description was included instead in a separate table from the same package [@citesubwaydata].
+Using `merge` from dplyr [@citedplyr], the incident was mapped to a generalized version of the description to match with the same set of generalized descriptions in the bus delay data.
+
+Similarly, "Line" was reported using acronyms instead of the full name, so for readability, the full name of the line which experienced the delay is now reported.
+
+### Issues with Dataset
+
+The vast majority of entries within the subway delay dataset had a recorded delay of 0 minutes.
+It seems as if a delay of 0 minutes really meant a delay of strictly less than 1 minute which is reasonable for a subway system.
+However, since this is an assumption, I have provided the filtered dataset for reference in the source repository, which only contains delay times greater than 0 minutes.
+
+Note that this dataset has been reduced to only 8,236 entries, comprising only 36.66% of the original dataset.
+
 ## Results
+
+### Delays as a function of Time
+
+```{r}
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-vstime
+#| fig-cap: "The amount of delays with respect to the time of day"
+#| fig-subcap: ["Bus Delays", "Subway Delays"]
+
+bus_delay_data |>
+  ggplot(aes(x=time)) +
+  geom_histogram(fill="blue", alpha=0.75) +
+  scale_x_time(breaks=scales::pretty_breaks(n=12)) +
+  theme(axis.text.x=element_text(angle=60, hjust=1))
+
+subway_delay_data |>
+  ggplot(aes(x=time)) +
+  geom_histogram(fill="blue", alpha=0.75) +
+  scale_x_time(breaks=scales::pretty_breaks(n=12)) +
+  theme(axis.text.x=element_text(angle=60, hjust=1))
+
+```
+
+```{r}
+#| echo: false
+#| message: false
+#| warning: false
+
+# Display plots
+
+```
 

### 2024-01-23 scripts/03-parse_codes.R
Better specify groupings

@@ -61,10 +61,15 @@ cleaned_subway_codes <-
   mutate(
     incident = case_when(
       startsWith(code, "E") ~ "Equipment/Mechanical",
-      startsWith(code, "M") ~ "Miscellaneous",
+      startsWith(code, "MUI") ~ "Security/Safety",
+      startsWith(code, "MUS") ~ "Security/Safety",
+      startsWith(code, "MUP") ~ "Security/Safety",
+      startsWith(code, "MUD") ~ "Equipment/Mechanical",
+      startsWith(code, "MUE") ~ "Equipment/Mechanical",
       startsWith(code, "P") ~ "Equipment/Mechanical",
       startsWith(code, "S") ~ "Security/Safety",
-      startsWith(code, "T") ~ "Operator"
+      startsWith(code, "T") ~ "Operator",
+      .default = "Miscellaneous"
     )
   )
 

### 2024-01-23 scripts/04-summarise_data.R

@@ -38,12 +38,6 @@ total_bus_delay_time_by_date <-
     mean_delay_time = (total_delay_time / n),
     .groups = "keep")
 
-# Summarize into average delay time per day of the week
-avg_bus_delay_time_by_day <-
-  total_bus_delay_time_by_date |>
-  group_by(day) |>
-  summarise(mean = mean(total_delay_time))
-
 # Summarize into the total number of delays per incident type (e.g. Equipment/Mechanical)
 total_num_bus_delays_by_incident <-
   bus_delay_data |>
@@ -78,12 +72,6 @@ total_subway_delay_time_by_date <-
     mean_delay_time = (total_delay_time / n),
     .groups = "keep")
 
-# Summarize into average delay time per day of the week
-avg_subway_delay_time_by_day <-
-  total_subway_delay_time_by_date |>
-  group_by(day) |>
-  summarise(mean = mean(total_delay_time))
-
 # Summarize into the total number of delays per incident type (e.g. Equipment/Mechanical)
 total_num_subway_delays_by_incident <-
   subway_delay_data |>
@@ -97,8 +85,6 @@ total_num_subway_delays_by_line <-
   summarise(n = n())
 
 # Save all the data
-write_csv(avg_bus_delay_time_by_day, "outputs/data/summaries/avg_bus_delay_time_by_day.csv")
-write_csv(avg_subway_delay_time_by_day, "outputs/data/summaries/avg_subway_delay_time_by_day.csv")
 
 write_csv(avg_num_bus_delays_by_day, "outputs/data/summaries/avg_num_bus_delays_by_day.csv")
 write_csv(avg_num_subway_delays_by_day, "outputs/data/summaries/avg_num_subway_delays_by_day.csv")
@@ -113,10 +99,8 @@ write_csv(total_num_subway_delays_by_line, "outputs/data/summaries/total_num_sub
 
 # Clean up workspace
 rm(list = c(
-  "avg_bus_delay_time_by_day",
   "avg_num_bus_delays_by_day",
   "avg_num_subway_delays_by_day",
-  "avg_subway_delay_time_by_day",
   "bus_delay_data",
   "subway_delay_data",
   "total_bus_delay_time_by_date",

### 2024-01-23 outputs/paper.qmd
Finish Paper

@@
 title: "Analyzing TTC Delay Data to Improve Future Efficiency"
-abstract: "TTC delays have been increasing over time, and especially now in a period of overhaul in many parts of the system, it becomes important to analyze the effects and relations subways have on bus delays and commuter loads.
-Using data published by the TTC and accessed through `opendatatoronto` we find that the greatest causes of delays are date and time, factors of which are out of anyones control, but must be accounted for to improve future operational efficieny.
-Further analysis must be made about the correlations with subways and buses to make a stronger argument."
+abstract: "TTC delays have been increasing over time, and especially now in a period of overhaul in many parts of the system, it becomes important to analyze the effects of subway and bus delays in relation with each other as well as commuter loads.
+Using data published by the TTC and accessed through `opendatatoronto` we find that the greatest causes of delays are date and time, and equipment quality.
+Date and time factors are out of anyones control, but must be accounted for to improve future operational efficieny, while equipment quality has also been shown to be important, with mechanical failures heavily decreasing the reliability of buses."
 thanks: "Code and data supporting this analysis is available at: https://github.com/prajogt/ttc_analysis.git"
@@
 So much so that leaving early is leaving on time.
-
-With the delays getting worse and service becoming more inconsistent [@TorontoStarTTCDelays], it is important to analyze potential causes and therefore enable the TTC to explore potential solutions.
-One of such factors is the tight coupling of the subway and bus systems.
+With the delays getting worse and service becoming more inconsistent [@TorontoStarTTCDelays], it is important to analyze potential causes, which would therefore enable the TTC to explore potential solutions.
+One of such possible factors is the tight coupling of the subway and bus systems.
 When issues occur with the subway, the overflow is sent to buses, either from detoured commuter traffic, or resource reallocation in the form of converting regular route buses to shuttle buses [@TorontoObserverShuttleBuses].
+Whether you, the reader, may be a commuter or not, having a good and reliable source of public transportation is crucial to reducing travel time for all city goers, not to mention its additional benefits for the environment.
 
 Within this paper we will analyze the correlations between the delays occurring on the subway and the delays that occur for buses.
-In particular, the correlation between mechanical or security issues on the subway, which tend to shut down a portion of the track, and with diversion and other external issues (here after labeled "Miscellaneous") that occur for buses.
-A logical correlation was made above, and a correlation shown in data would go to further a cause and effect relationship.
-
-Later we will see the general affects of the commuter load on the transit system and how the system can be better optimized and fail-safed to prevent such frequent delays.
+In particular, the correlation between mechanical or security issues on the subway, which tend to shut down a portion of the track, and similar equipment and now operator failures that occur for buses.
+We will see the general affects of the commuter load on the transit system and how the system can be better optimized or at least supported with redundancies to prevent such frequent delays.
 The paper will show the trends of the amount of delays for both systems as a function of day and time.
 It was found that work days put additional load on both systems, the total amount of delays during weekdays being significantly higher than on other days.
+It is found that the cycle of the traditional work week and work day, the environment in which each exists in with regards to the facilities, equipment, and workers, have a strong correlation with delays in the TTC.
 
-The datasets used for this analysis are described in the Data section, as well as the process that was done to clean the data into comparable datasets.
+The next section will cover the content of the data and analyses on the results that were obtained from the datasets.
 
@@
 
-Data was downloaded, cleaned, parsed, analyzed, and visualized using R [@citeR], a statistical programming language, with package support from `tidyverse` [@citetidyverse], a collection of libraries which included the following packages that were utilized: \* `ggplot2` [@citeggplot] \* `dplyr` [@citedplyr] \* `readr` [@citereadr] \* `tibble` [@citetibble]
+Data was downloaded, cleaned, parsed, analyzed, and visualized using R [@citeR], a statistical programming language, with package support from `tidyverse` [@citetidyverse], a collection of libraries which included the following packages that were utilized:
+
+* `ggplot2` [@citeggplot] 
+* `dplyr` [@citedplyr] 
+* `readr` [@citereadr] 
+* `tibble` [@citetibble]
 
@@
 
-For @tbl-subway_delay_data the additional parameter "Line" represents the subway line in which the delay took place in. For those not familiar
+For @tbl-subway_delay_data the additional parameter "Line" represents the subway line in which the delay took place in.
+For those not familiar, these are the names of the different subway tracks the TTC utilizes.
 
-In total @tbl-bus_delay_data contains 51,406 entries, whilst @tbl-subway_delay_data contains 22,465 entries.
+In total @tbl-bus_delay_data contains 51,406 entries, whilst @tbl-subway_delay_data contains 22,465 entries, a significant difference.
 
@@
 Using `merge` from dplyr [@citedplyr], the incident was mapped to a generalized version of the description to match with the same set of generalized descriptions in the bus delay data.
+This selection into sections based on the description important to note due to the potential subjectivity of the sorting.
+There are incident descriptions in which the record may fall in one or the other.
+Please see the data cleaning scripts for more details.
 
-Similarly, "Line" was reported using acronyms instead of the full name, so for readability, the full name of the line which experienced the delay is now reported.
-
-### Issues with Dataset
+#### Issues with Dataset {#sec-issues_with_dataset}
 
@@
 
-Note that this dataset has been reduced to only 8,236 entries, comprising only 36.66% of the original dataset.
+Note that this filtered dataset has been reduced to only 8,236 entries, comprising only 36.66% of the original dataset.
 
-## Results
-
-### Delays as a function of Time
+### By Time
 
 ```{r}
+#| eval: true
 #| echo: false
@@
 #| fig-subcap: ["Bus Delays", "Subway Delays"]
+#| layout-ncol: 2
+
+# Display the total amount of delays (on any date) for
+# a particular time
 
 bus_delay_data |>
-  ggplot(aes(x=time)) +
-  geom_histogram(fill="blue", alpha=0.75) +
-  scale_x_time(breaks=scales::pretty_breaks(n=12)) +
-  theme(axis.text.x=element_text(angle=60, hjust=1))
+  ggplot(aes(x = time)) +
+  geom_histogram(fill = "blue", alpha = 0.75) +
+  scale_x_time(breaks = scales::pretty_breaks(n = 12)) +
+  theme_light() +
+  theme(axis.text.x = element_text(angle = 60, hjust = 1))
 
@@
   ggplot(aes(x=time)) +
-  geom_histogram(fill="blue", alpha=0.75) +
-  scale_x_time(breaks=scales::pretty_breaks(n=12)) +
-  theme(axis.text.x=element_text(angle=60, hjust=1))
+  geom_histogram(fill = "red", alpha = 0.75) +
+  scale_x_time(breaks = scales::pretty_breaks(n = 12)) +
+  theme_light() + 
+  theme(axis.text.x = element_text(angle = 60, hjust = 1))
+
+```
+
+@fig-vsttime shows the difference in the trends of delays over time.
+Whilst bus delays heavily spike during rush hours where traffic alongside the buses has an impact on their operation, subway delays show a less significant increase in delays.
+Something important to note is the difference in scale that is present for all the graphs presented in this report, with the peak count of delays for @fig-vstime-2 being only around half of the peak count of delays for @fig-vstime-1.
+
+Regardless of the scale / degree of effect that the work cycle has on the TTC, it is evident that there is a strong correlation between the traditional schedule of the work day has with the amount of delays that occur.
+In a similar manner, the work week cycle has a similar correlation.
+
+### By Day
+
+```{r}
+#| eval: true
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-avg_num_vsday
+#| fig-cap: "The average number of delays occurred on a given day of the week"
+#| fig-subcap: ["Bus Delays", "Subway Delays"]
+#| layout-ncol: 2
+
+# Load in summarized data
+avg_num_bus_delays_by_day <- read_csv(
+   file = "data/summaries/avg_num_bus_delays_by_day.csv",
+   show_col_types = FALSE
+)
+
+avg_num_bus_delays_by_day$day <- 
+  factor(avg_num_bus_delays_by_day$day, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
+
+avg_num_bus_delays_by_day |>
+  ggplot(aes(x = day, y = mean_num_delays, fill = day)) +
+  geom_bar(stat = 'identity') +
+  xlab("Day of the Week") +
+  ylab("Mean Number of Delays") + 
+  theme_light() +
+  theme(axis.text.x = element_text(angle = 60, hjust = 1))
+
+avg_num_subway_delays_by_day <- read_csv(
+   file = "data/summaries/avg_num_subway_delays_by_day.csv",
+   show_col_types = FALSE
+)
+
+avg_num_subway_delays_by_day$day <- 
+  factor(avg_num_subway_delays_by_day$day, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
+
+avg_num_subway_delays_by_day |>
+  ggplot(aes(x = day, y = mean_num_delays, fill = day)) +
+  geom_bar(stat = 'identity') +
+  xlab("Day of the Week") +
+  ylab("Mean Number of Delays") + 
+  theme_light() +
+  theme(axis.text.x = element_text(angle = 60, hjust = 1))
+
+```
+
+@fig-avg_num_vsday shows the similar trends that are present in the average number of delays that occur on certain days of the week.
+On average Sundays in 2023 produce the lowest amount of delays.
+Buses and subways average 101.00 and 50.75 delays respectively on Sundays.
+It seems that the added commuter load or the assumption of an increased commuter load causes more delays.
+We cannot say for certain which of the two it is, as it is equally likely that the reason that more delays occur on certain days is because more buses and trains are scheduled more frequently to accommodate the load of working commuters.
+Equally, it is possible that the increase in delays is caused by a human actor.
+As shown by @fig-vsincident, a large part of the delays for subways are security and safety issues which is almost entirely comprised of human effects.
+What we can say now however, is that more preparation and further analysis into commuter effects must be done by the TTC to lower the large amount of delays caused.
+
+```{r}
+#| eval: true
+#| echo: false
+#| message: false
+#| warning: false
+#| label: tbl-totalbus
+#| tbl-cap: "The total and average time in minutes wasted on delays"
+
+total_bus_delay_time_by_date <- 
+  read_csv(
+   file = "data/summaries/total_bus_delay_time_by_date.csv",
+   show_col_types = FALSE
+)
+
+# Take 5 random rows from the summarized data 
+set.seed(302)
+total_bus_delay_time_by_date[sample(nrow(total_bus_delay_time_by_date), 5), ] |>
+  kable(
+    col.names = c("Date", "Day", "Total Delay Time", "Delay Count", "Average Delay Time"),
+    booktabs = TRUE
+  )
 
@@
 ```{r}
+#| eval: true
 #| echo: false
@@
 #| warning: false
+#| label: fig-fullyear
+#| fig-cap: "Bus Total and Average Delays in Time (Minutes)"
+#| fig-subcap: ["Total Delay Time", "Average Delay Time"]
+#| layout-ncol: 2
 
-# Display plots
+total_bus_delay_time_by_date |>
+  ggplot(aes(x = date, y = total_delay_time)) + 
+  geom_point() +
+  geom_smooth(method = "loess") +
+  theme_light()
+
+total_bus_delay_time_by_date |>
+  ggplot(aes(x = date, y = mean_delay_time)) + 
+  geom_point() +
+  geom_smooth(method = "loess") +
+  theme_light()
 
@@
 
-## Discussion
+Here @tbl-totalbus is a random sampling of dates in 2023 and the total amount both in time (minutes) and quantity of delays that occur on that particular date.
+So much time is wasted on delays for commuters, supporting the claim that City News [@CityNewsBGrade] made about the quality of service.
+Seeing this further in @fig-fullyear, we see that the average bus commuter spends around 20-25 minutes every day waiting for delays, and the TTC overall has wasted around 30,000 minutes (500 hours) of commuters time as a whole.
+
+### By Incident
+
+```{r}
+#| eval: true
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-vsincident
+#| fig-cap: "The average number of delays occurred due to incident types"
+#| fig-subcap: ["Bus Delays", "Subway Delays"]
+#| layout-ncol: 2
+
+# Load summarized data about number of delays vs incidents
+
+bus_delays_by_incident <- read_csv(
+   file = "data/summaries/total_num_bus_delays_by_incident.csv",
+   show_col_types = FALSE
+) 
+bus_delays_by_incident |> 
+  ggplot(aes(x=incident, y=n, fill=incident)) +
+  geom_bar(stat = "identity") +
+  xlab("Incident Type") +
+  ylab("Delay Count") +
+  theme_light() +
+  theme(axis.text.x = element_text(angle = 60, hjust = 1))
+
+subway_delays_by_incident <- read_csv(
+   file = "data/summaries/total_num_subway_delays_by_incident.csv",
+   show_col_types = FALSE
+) 
+subway_delays_by_incident |> 
+  ggplot(aes(x = incident, y = n, fill = incident)) +
+  geom_bar(stat = "identity") +
+  xlab("Incident Type") +
+  ylab("Delay Count") +
+  theme_light() +
+  theme(axis.text.x = element_text(angle = 60, hjust = 1))
+
+```
+
+From @fig-vsincident-2 we can see that the leading cause of delay on the subways is "Security/Safety" incidents with 15,367 delays caused by such issues, with the next leading incident being "Equipment/Mechanical" issues with 4,005 delays occurred.
+"Security/Safety" covers track invasions, unsanitary trains, fire, etc.
+and "Equipment/Mechanical" covers any problems occurring with the functionality of the trains, the station, and the tracks.
+From personal experience, these results are expected.
+Comparatively, bus delays in @fig-vsincident-1 are majorly caused by "Equipment/Mechanical" issues with 18,786 caused.
+The second leading incident type is operator issues which covers any problems with the driver of the buses.
+These incidents account for 14,400 delays.
+These results pose an interesting connection.
+
+From the TTC website, we know that they have been moving towards implementing automation in the operation of subway services since 2022[@TTCAutomation].
+This has not been possible for buses which is why operator error is still high on the list.
+This could also be explained by the ratio of operators to commuters.
+Whilst each bus requires an operator, multiple cars on a subway only require 1-2 operators especially with the addition of automated signalling.
+Similarly, the lack of employee oversight within stations and subways can explain the difference in security and safety issues.
+Buses don't have much to oversee, however since stations are large and employees are only stationed within the entrance floor, prevention of disruptive behaviour is more commonplace as displayed by @fig-vsincident-2 above.
+
+### By Line
+
+```{r}
+#| eval: true
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-vsline
+#| fig-cap: "The number of delays on each TTC Line"
+
+subway_delays_by_line <- read_csv(
+   file = "data/summaries/total_num_subway_delays_by_line.csv",
+   show_col_types = FALSE
+)
+
+subway_delays_by_line |> 
+  ggplot(aes(x = line, y = n, fill = line)) +
+  geom_bar(stat = "identity") + 
+  theme_light() +
+  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
+  xlab("TTC Line") +
+  ylab("Number of Delays")
+
+```
+From @fig-vsline we see an expected result, where the longest (and oldest) lines with Yonge-University having 38 stations and Bloor-Danforth having 31 compared to Sheppard's 5 and Scarborough's 6 (previously existing) stations.
+With the load that these older lines are experiencing on a daily basis, the result from our analysis on delay causes is further supported.
+Constant maintenance and repair is required for these lines, and increased ridership also puts extra stress on the systems.
+Further engineering analysis is required to definitively know the extent of the effects of stress loads on these lines, and if those are the cause for the increase in delays.
 
@@
 
+In conclusion, the analysis of TTC delays has revealed critical factors impacting the efficiency and reliability of Toronto's major transit system.
+Analyzing data obtained from the TTC through `opendatatoronto`, we identified date and time, equipment quality, and operation failure as the major contributors to delays.
+Although we cannot control date and time factors, acknowledging them and their effects on the system is vital for improving future operational efficiency.
+Recognizing the impact of equipment quality, particularly in relation to bus mechanical failures, highlights the need for specific upgrades and maintenance in transit infrastructure.
+The insights from this analysis can guide future spending and research, in an attempt to be a part of the effort to enhance the TTC system for the benefit of both commuters and the city.
+
 ## Reference

### 2024-01-23 scripts/04-summarise_data.R
Remove uneccessary library

@@ -8,7 +8,6 @@
 
 #### Workspace setup ####
 library(tidyverse)
-library(lubridate)
 
 # Read in bus data
 bus_delay_data <- 

### 2024-01-23 scripts/04-summarise_data.R
Add prereq

@@ -4,7 +4,7 @@
 # Date: 23 January 2024
 # Contact: tim.prajogi@mail.utoronto.ca
 # License: MIT
-# Prerequisites: 01-download_data, 02-clean_data
+# Prerequisites: 01-download_data, 02-clean_data, 03-parse_codes
 
 #### Workspace setup ####
 library(tidyverse)

### 2024-01-25 outputs/paper.qmd
Fix typo
Update paper per peer reviews

@@
 number-sections: true
-bibliography: refrences.bib
+bibliography: references.bib
+header-includes:
+  \usepackage{placeins}
 ---
@@
 The main source public transportation of public transportation in the city of Toronto is the TTC (Toronto Transit Commission).
-The most common methods of this public transportation is subways and buses.
-For people who frequently use these methods, delays are common and expected [@CityNewsBGrade].
-So much so that leaving early is leaving on time.
+From this commission, the most common methods used are subways and buses.
+For people who frequently use these methods, delays are common and expected [@CityNewsBGrade], so much so that leaving early is leaving on time.
 With the delays getting worse and service becoming more inconsistent [@TorontoStarTTCDelays], it is important to analyze potential causes, which would therefore enable the TTC to explore potential solutions.
@@
 When issues occur with the subway, the overflow is sent to buses, either from detoured commuter traffic, or resource reallocation in the form of converting regular route buses to shuttle buses [@TorontoObserverShuttleBuses].
-Whether you, the reader, may be a commuter or not, having a good and reliable source of public transportation is crucial to reducing travel time for all city goers, not to mention its additional benefits for the environment.
+Whether one uses these methods or not, having a good and reliable source of public transportation is crucial to reducing travel time for all city goers, not to mention its additional benefits for the environment.
 
@@
 
-The datasets used for this report are the latest delay statistics from the 2023 year, `TTC SUbway Delay Data` [@citesubwaydata] and `TTC Bus Delay Data` [@citebusdata], both of which were published by the TTC.
+The datasets used for this report are the latest delay statistics from the 2023 year, `TTC Subway Delay Data` [@citesubwaydata] and `TTC Bus Delay Data` [@citebusdata], both of which were published by the TTC.
 
-Data was downloaded, cleaned, parsed, analyzed, and visualized using R [@citeR], a statistical programming language, with package support from `tidyverse` [@citetidyverse], a collection of libraries which included the following packages that were utilized:
+These delay statistics were downloaded, cleaned, parsed, analyzed, and visualized using R [@citeR], a statistical programming language, with package support from `tidyverse` [@citetidyverse], a collection of libraries which included the following packages that were utilized:
 
@@
 
-Both @tbl-bus_delay_data and @tbl-subway_delay_data contain columns "Date" and "Time" in which the delay had occurred, and an "Incident" in which the type of incident that had occurred out of these 4 general types: "Operator" (for driver/operator related issues), "Security/Safety" (for fire, police, trespassing, etc.), "Equipment/Mechanical" (for equipment, track, vehicle, etc., failures), and "Miscellaneous" (for diversions and other external issues).
+Both the bus and subway delay dataset contain columns "Date" and "Time" in which the delay had occurred, and an "Incident" in which the type of incident that had occurred out of these 4 general types: "Operator" (for driver/operator related issues), "Security/Safety" (for fire, police, trespassing, etc.), "Equipment/Mechanical" (for equipment, track, vehicle, etc., failures), and "Miscellaneous" (for diversions and other external issues).
 
-For @tbl-bus_delay_data and @tbl-subway_delay_data the "Min Delay" and "Min Gap" are in minutes.
+"Min Delay" and "Min Gap" are represented in minutes.
 Where "Min Delay" represents the delay in the schedule for the next bus and "Min Gap" is the scheduled time from the bus ahead of the following bus.
 
-For @tbl-subway_delay_data the additional parameter "Line" represents the subway line in which the delay took place in.
+For the subway delay dataset, the additional feature "Line" represents the subway line in which the delay took place in.
 For those not familiar, these are the names of the different subway tracks the TTC utilizes.
+The 4 TTC lines operating in 2023 are labeled with their official name, however, the dataset also includes incidents that occurred at intersections between Yonge-University and Bloor-Danforth, which for the purposes of this report was considered separately.
+Additionally, lines which did not match any of the existing TTC lines were labeled "Other", as if they were significant enough to include in this dataset, they should have a relationship with the subway Incidents.
 
-In total @tbl-bus_delay_data contains 51,406 entries, whilst @tbl-subway_delay_data contains 22,465 entries, a significant difference.
+In total the bus delay dataset that @tbl-bus_delay_data samples contains 51,406 entries, whilst the subway delay dataset that @tbl-subway_delay_data samples contains 22,465 entries, a significant difference.
+To be precise, the amount of delays that have occurred in 2023 is 2.29 times the amount of delays that have occurred for the subway.
+If we further consider the filtered data, the difference amount of significant (in this case meaning with a delay time greater than 1 minute) is widened as will be discussed in @sec-issues_with_dataset.
 
@@
 #| warning: false
-#| label: fig-vstime
+#| label: fig-vstime1
 #| fig-cap: "The amount of delays with respect to the time of day"
-#| fig-subcap: ["Bus Delays", "Subway Delays"]
-#| layout-ncol: 2
+#| fig-width: 4
 
@@
   scale_x_time(breaks = scales::pretty_breaks(n = 12)) +
+  xlab("Time") +
+  ylab("Count") +
   theme_light() +
-  theme(axis.text.x = element_text(angle = 60, hjust = 1))
+  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
+  theme(axis.text=element_text(size=12),
+      axis.title=element_text(size=14,face="bold"))
+
+```
+
+```{r}
+#| eval: true
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-vstime2
+#| fig-cap: "The amount of delays with respect to the time of day"
+#| fig-width: 4
+
+# Display the total amount of delays (on any date) for
+# a particular time
 
@@
   scale_x_time(breaks = scales::pretty_breaks(n = 12)) +
+  xlab("Time") +
+  ylab("Count") +
   theme_light() + 
-  theme(axis.text.x = element_text(angle = 60, hjust = 1))
+  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
+  theme(axis.text=element_text(size=12),
+        axis.title=element_text(size=14,face="bold"))
 
@@
 
-@fig-vsttime shows the difference in the trends of delays over time.
+@fig-vstime1 and @fig-vstime2 shows the difference in the trends of delays over time.
 Whilst bus delays heavily spike during rush hours where traffic alongside the buses has an impact on their operation, subway delays show a less significant increase in delays.
-Something important to note is the difference in scale that is present for all the graphs presented in this report, with the peak count of delays for @fig-vstime-2 being only around half of the peak count of delays for @fig-vstime-1.
+Subways do not have to deal with external traffic factors which is likely the cause of buses showing a larger amount of delays during those hours in relation with other times.
+Something important to note is the difference in scale that is present for all the graphs presented in this report, with the peak count of delays for @fig-vstime2 being only around half of the peak count of delays for @fig-vstime1.
 
@@
 Equally, it is possible that the increase in delays is caused by a human actor.
-As shown by @fig-vsincident, a large part of the delays for subways are security and safety issues which is almost entirely comprised of human effects.
+As shown by @fig-vsincident1 and @fig-vsincident2, a large part of the delays for subways are security and safety issues which is almost entirely comprised of human effects.
 What we can say now however, is that more preparation and further analysis into commuter effects must be done by the TTC to lower the large amount of delays caused.
@@
 #| warning: false
-#| label: fig-vsincident
+#| label: fig-vsincident1
 #| fig-cap: "The average number of delays occurred due to incident types"
-#| fig-subcap: ["Bus Delays", "Subway Delays"]
-#| layout-ncol: 2
 
@@
   theme_light() +
-  theme(axis.text.x = element_text(angle = 60, hjust = 1))
+  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
+  guides(fill=guide_legend(title="Incident Type"))
+
+```
+
+```{r}
+#| eval: true
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-vsincident2
+#| fig-cap: "The average number of delays occurred due to incident types"
 
@@
   theme_light() +
-  theme(axis.text.x = element_text(angle = 60, hjust = 1))
+  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
+  guides(fill=guide_legend(title="Incident Type"))
 
@@
 
-From @fig-vsincident-2 we can see that the leading cause of delay on the subways is "Security/Safety" incidents with 15,367 delays caused by such issues, with the next leading incident being "Equipment/Mechanical" issues with 4,005 delays occurred.
+From @fig-vsincident2 we can see that the leading cause of delay on the subways is "Security/Safety" incidents with 15,367 delays caused by such issues, with the next leading incident being "Equipment/Mechanical" issues with 4,005 delays occurred.
 "Security/Safety" covers track invasions, unsanitary trains, fire, etc.
@@
 From personal experience, these results are expected.
-Comparatively, bus delays in @fig-vsincident-1 are majorly caused by "Equipment/Mechanical" issues with 18,786 caused.
+Comparatively, bus delays in @fig-vsincident1 are majorly caused by "Equipment/Mechanical" issues with 18,786 caused.
 The second leading incident type is operator issues which covers any problems with the driver of the buses.
@@
 Similarly, the lack of employee oversight within stations and subways can explain the difference in security and safety issues.
-Buses don't have much to oversee, however since stations are large and employees are only stationed within the entrance floor, prevention of disruptive behaviour is more commonplace as displayed by @fig-vsincident-2 above.
+Buses don't have much to oversee, however since stations are large and employees are only stationed within the entrance floor, prevention of disruptive behaviour is more commonplace as displayed by @fig-vsincident2 above.
 
-### By Line
+### By Line {#sec-by_line}
 
@@
 ```
+
 From @fig-vsline we see an expected result, where the longest (and oldest) lines with Yonge-University having 38 stations and Bloor-Danforth having 31 compared to Sheppard's 5 and Scarborough's 6 (previously existing) stations.
@@
 
+\FloatBarrier
 ## Conclusion
@@
 
-## Reference
+\newpage
+## References

### 2024-01-25 outputs/paper.qmd
Update figures

@@
 #| fig-width: 4
+#| fig-height: 3
 
@@
   theme_light() +
-  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
-  theme(axis.text=element_text(size=12),
-      axis.title=element_text(size=14,face="bold"))
+  theme(axis.text.x = element_text(angle = 60, hjust = 1))
 
@@
   theme_light() + 
-  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
-  theme(axis.text=element_text(size=12),
-        axis.title=element_text(size=14,face="bold"))
+  theme(axis.text.x = element_text(angle = 60, hjust = 1))
 
@@
   geom_smooth(method = "loess") +
-  theme_light()
+  theme_light() +
+  xlab("Date") +
+  ylab("Total Delay Time (Minutes)")
 

### 2025-06-19 scripts/00-simulate_data.R
updated scripts to python

@@ -1,245 +1,0 @@
-#### Preamble ####
-# Purpose: Simulate delay data
-# Author: Timothius Prajogi
-# Date: 23 January 2024
-# Contact: tim.prajogi@mail.utoronto.ca
-# License: MIT
-# Prerequisites: None
-
-
-#### Workspace setup ####
-library(tidyverse)
-
-
-#### Preamble ####
-# Purpose: Simulate delay data
-# Author: Timothius Prajogi
-# Date: 23 January 2024
-# Contact: tim.prajogi@mail.utoronto.ca
-# License: MIT
-# Prerequisites: None
-
-
-#### Workspace setup ####
-library(tidyverse)
-set.seed(302)
-
-## Simulated situation attributes ##
-
-# The possible incident types causing the delay
-sim_incident_types <- 
-  c(
-    "Equipment/Mechanical", 
-    "Miscellaneous", 
-    "Security/Safety", 
-    "Operator"
-  )
-
-# Expected distribution of incident types
-sim_incident_probabilities <-
-  c(
-    0.3,
-    0.2,
-    0.4,
-    0.2
-  )
-
-# The days of the week the buses and subways operate
-sim_operating_days <-
-  c(
-    "Sunday",
-    "Monday",
-    "Tuesday",
-    "Wednesday",
-    "Thursday",
-    "Friday",
-    "Saturday"
-  )
-
-# Assumption that delays change depending on the day
-sim_probability_of_subway_delay <-
-  c(
-    0.18, # Sort of high on Sundays
-    0.14, # Slightly lower as people go back to school/work
-    0.10, # Low on Tuesday
-    0.10, # Low on Wednesday
-    0.11, # Low on Thursday
-    0.17, # High on Friday
-    0.20  # High on Saturday
-  )
-
-# Sample 1 time from a possion dist with an average
-# of 20,000 delays in 1 day
-num_subway_delays <- rpois(1, 20000)
-
-# Assumption that there are more bus delays (strong base since bus delay 
-# dataset is twice the size) and subway delays and bus delays are correlated
-num_bus_delays <- 1.5 * num_subway_delays
-
-
-## Simulated situation attributes ##
-
-# The possible incident types causing the delay
-sim_incident_types <- 
-  c(
-    "Equipment/Mechanical", 
-    "Miscellaneous", 
-    "Security/Safety", 
-    "Operator"
-  )
-
-# Expected distribution of incident types
[…]

### 2025-06-19 scripts/00-simulate_data.py

@@ -1,0 +1,178 @@
+#### Preamble ####
+# Purpose: Simulate delay data
+# Author: Timothius Prajogi
+# Date: 23 January 2024
+# Contact: tim.prajogi@mail.utoronto.ca
+# License: MIT
+# Prerequisites: None
+
+
+#### Workspace setup ####
+import pandas as pd
+import numpy as np
+import matplotlib.pyplot as plt
+
+np.random.seed(302)
+
+
+## Simulated situation attributes ##
+
+# The possible incident types causing the delay
+sim_incident_types = [
+    "Equipment/Mechanical",
+    "Miscellaneous",
+    "Security/Safety",
+    "Operator"
+]
+
+# Expected distribution of incident types
+sim_incident_probabilities = [
+    0.3,
+    0.2,
+    0.4,
+    0.2
+]
+
+# The days of the week the buses and subways operate
+sim_operating_days = [
+    "Sunday",
+    "Monday",
+    "Tuesday",
+    "Wednesday",
+    "Thursday",
+    "Friday",
+    "Saturday"
+]
+
+# Assumption that delays change depending on the day
+sim_probability_of_subway_delay = [
+    0.18,  # Sort of high on Sundays
+    0.14,  # Slightly lower as people go back to school/work
+    0.10,  # Low on Tuesday
+    0.10,  # Low on Wednesday
+    0.11,  # Low on Thursday
+    0.17,  # High on Friday
+    0.20   # High on Saturday
+]
+
+# Sample 1 time from a possion dist with an average
+# of 20,000 delays in 1 day
+num_subway_delays = np.random.poisson(20000)
+
+# Assumption that there are more bus delays (strong base since bus delay 
+# dataset is twice the size) and subway delays and bus delays are correlated
+num_bus_delays = int(1.5 * num_subway_delays)
+
+
+## Simulated situation attributes ##
+
+# The possible incident types causing the delay
+sim_incident_types = [
+    "Equipment/Mechanical",
+    "Miscellaneous",
+    "Security/Safety",
+    "Operator"
+]
+
+# Expected distribution of incident types
+sim_incident_probabilities = [
+    0.3,
+    0.2,
+    0.4,
+    0.2
+]
+
+# The days of the week the buses and subways operate
+sim_operating_days = [
+    "Sunday",
+    "Monday",
+    "Tuesday",
+    "Wednesday",
+    "Thursday",
+    "Friday",
+    "Saturday"
+]
+
[…]

### 2025-06-19 scripts/01-download_data.R

@@ -1,55 +1,0 @@
-#### Preamble ####
-# Purpose: Download TTC delay data from opendatatoronto
-# Author: Timothius Prajogi
-# Date: 23 January 2024
-# Contact: tim.prajogi@mail.utoronto.ca
-# License: MIT
-# Prerequisites: none
-
-#### Workplace setup ####
-
-library(opendatatoronto)
-library(tidyverse)
-
-#### Download and save TTC data ####
-
-## Subway delay data
-## -----------------
-## 996cfe8d-fb35-40ce-b569-698d51fc683b -> Package
-## 2fbec48b-33d9-4897-a572-96c9f002d66a -> Delay Statistics Resource 
-## 3900e649-f31e-4b79-9f20-4731bbfd94f7 -> Delay Code Meaning Resource
-
-raw_subway_data <- 
-  list_package_resources("996cfe8d-fb35-40ce-b569-698d51fc683b") |> 
-  filter (id == "2fbec48b-33d9-4897-a572-96c9f002d66a") |> 
-  get_resource() 
-
-write_csv(raw_subway_data, 'inputs/data/raw_subway_delay_statistics.csv')
-
-subway_delay_codes <- 
-  list_package_resources("996cfe8d-fb35-40ce-b569-698d51fc683b") |> 
-  filter (id == "3900e649-f31e-4b79-9f20-4731bbfd94f7") |> 
-  get_resource()
-
-write_csv(subway_delay_codes, 'inputs/data/raw_subway_delay_codes.csv')
-
-## Bus delay data
-## --------------
-## e271cdae-8788-4980-96ce-6a5c95bc6618 -> Package
-## 10802a64-9ac0-4f2e-9538-04800a399d1e -> Resource
-
-raw_bus_data <-
-  list_package_resources("e271cdae-8788-4980-96ce-6a5c95bc6618") |> 
-  filter (id == "10802a64-9ac0-4f2e-9538-04800a399d1e") |> 
-  get_resource() 
-
-write_csv(raw_bus_data, 'inputs/data/raw_bus_delay_statistics.csv')
-
-# Clean up workspace
-rm(list = c(
-  "raw_bus_data",
-  "raw_subway_data",
-  "subway_delay_codes"
-))
-
-

### 2025-06-19 scripts/01-download_data.py

@@ -1,0 +1,53 @@
+#### Preamble ####
+# Purpose: Download TTC delay data from opendatatoronto
+# Author: Timothius Prajogi
+# Date: 23 January 2024
+# Contact: tim.prajogi@mail.utoronto.ca
+# License: MIT
+# Prerequisites: none
+
+
+delay_stats_url = "https://ckan0.cf.opendata.inter.prod-toronto.ca/dataset/996cfe8d-fb35-40ce-b569-698d51fc683b/resource/3900e649-f31e-4b79-9f20-4731bbfd94f7/download/ttc-subway-delay-codes.xlsx"
+delay_code_url = "https://ckan0.cf.opendata.inter.prod-toronto.ca/dataset/996cfe8d-fb35-40ce-b569-698d51fc683b/resource/2fbec48b-33d9-4897-a572-96c9f002d66a/download/ttc-subway-delay-2023.xlsx"
+bus_delay_url = "https://ckan0.cf.opendata.inter.prod-toronto.ca/dataset/e271cdae-8788-4980-96ce-6a5c95bc6618/resource/10802a64-9ac0-4f2e-9538-04800a399d1e/download/ttc-bus-delay-data-2023.xlsx"
+
+#### Preamble ####
+# Purpose: Download TTC delay data from opendatatoronto
+# Author: Timothius Prajogi
+# Date: 23 January 2024
+# Contact: tim.prajogi@mail.utoronto.ca
+# License: MIT
+# Prerequisites: none
+
+#### Workplace setup ####
+
+import pandas as pd
+import requests
+from io import BytesIO
+
+#### Download and save TTC data ####
+
+# Provided direct download links
+delay_stats_url = "https://ckan0.cf.opendata.inter.prod-toronto.ca/dataset/996cfe8d-fb35-40ce-b569-698d51fc683b/resource/3900e649-f31e-4b79-9f20-4731bbfd94f7/download/ttc-subway-delay-codes.xlsx"
+delay_code_url = "https://ckan0.cf.opendata.inter.prod-toronto.ca/dataset/996cfe8d-fb35-40ce-b569-698d51fc683b/resource/2fbec48b-33d9-4897-a572-96c9f002d66a/download/ttc-subway-delay-2023.xlsx"
[…]

### 2025-06-19 scripts/02-clean_data.R

@@ -1,185 +1,0 @@
-#### Preamble ####
-# Purpose: Clean raw datasets
-# Author: Timothius Prajogi
-# Date: 23 January 2024
-# Contact: tim.prajogi@mail.utoronto.ca
-# License: MIT
-# Prerequisites: 01-download_data
-
-#### Workspace setup ####
-library(tidyverse)
-library(janitor)
-
-#### Read and clean raw data ####
-
-## Read in Bus data
-raw_bus_data <- 
-  read_csv(
-     file = "inputs/data/raw_bus_delay_statistics.csv",
-     show_col_types = FALSE
-  )
-
-# Clean names
-cleaned_bus_data <-
-  clean_names(raw_bus_data)
-
-# Select only relevant columns
-cleaned_bus_data <-
-  cleaned_bus_data |>
-  select(
-    date,
-    time,
-    day,
-    incident,
-    min_delay,
-    min_gap
-  )
-
-# Filter out situations in which the delay is 0, meaning the incident that
-# occurred didn't affect service
-cleaned_bus_data <-
-  cleaned_bus_data |>
-  filter(min_gap > 0)
-
-# Save cleaned bus data
-write_csv(cleaned_bus_data, 'inputs/data/bus_delay_statistics.csv')
-
-## SUBWAY DATA
-
-# Read in subway statistics
-raw_subway_data <-
-  read_csv(
-    file = "inputs/data/raw_subway_delay_statistics.csv",
-    show_col_types = FALSE
-  )
-
-# Clean names
-cleaned_subway_data <-
-  clean_names(raw_subway_data)
-
-# Select only relevant columns
-cleaned_subway_data <- 
-  cleaned_subway_data |>
-  select(
-    date,
-    time,
-    day,
-    code,
-    min_delay,
-    min_gap,
-    line # additional for subway (no similar grouping for buses)
-  )
-
-# Save cleaned subway data
-write_csv(cleaned_subway_data, 'inputs/data/subway_delay_statistics.csv')
-
-# The reason 2 different .csv files are stored is that the vast majority
-# of rows had a delay of 0 minutes. It seems as if a delay of 0 minutes
-# means a delay of < 1 minute which is reasonable for a subway system.
-# However, since this is an assumption, I also kept the filtered 
-# dataset for reference.
-
[…]

### 2025-06-19 scripts/02-clean_data.py

@@ -1,0 +1,139 @@
+#### Preamble ####
+# Purpose: Clean raw datasets
+# Author: Timothius Prajogi
+# Date: 23 January 2024
+# Contact: tim.prajogi@mail.utoronto.ca
+# License: MIT
+# Prerequisites: 01-download_data
+
+#### Workspace setup ####
+
+import pandas as pd
+
+#### Read and clean raw data ####
+
+## Read in Bus data
+raw_bus_data = pd.read_csv("inputs/data/raw_bus_delay_statistics.csv")
+
+# Clean names
+cleaned_bus_data = raw_bus_data.rename(columns=lambda x: x.strip().lower().replace(" ", "_"))
+
+# Select only relevant columns
+cleaned_bus_data = cleaned_bus_data[[
+    "date",
+    "time",
+    "day",
+    "incident",
+    "min_delay",
+    "min_gap"
+]]
+
+# Filter out situations in which the delay is 0, meaning the incident that
+# occurred didn't affect service
+cleaned_bus_data = cleaned_bus_data[cleaned_bus_data["min_gap"] > 0]
+
+# Save cleaned bus data
+cleaned_bus_data.to_csv("inputs/data/bus_delay_statistics.csv", index=False)
+
+
+## SUBWAY DATA
+
+# Read in subway statistics
+raw_subway_data = pd.read_csv("inputs/data/raw_subway_delay_statistics.csv")
+
+# Clean names
+cleaned_subway_data = raw_subway_data.rename(columns=lambda x: x.strip().lower().replace(" ", "_"))
+
+# Select only relevant columns
+cleaned_subway_data = cleaned_subway_data[[
+    "date",
+    "time",
+    "day",
+    "code",
+    "min_delay",
+    "min_gap",
+    "line"  # additional for subway (no similar grouping for buses)
+]]
+
+# Save cleaned subway data
+cleaned_subway_data.to_csv("inputs/data/subway_delay_statistics.csv", index=False)
+
+# The reason 2 different .csv files are stored is that the vast majority
+# of rows had a delay of 0 minutes. It seems as if a delay of 0 minutes
+# means a delay of < 1 minute which is reasonable for a subway system.
+# However, since this is an assumption, I also kept the filtered 
+# dataset for reference.
+
+# Filter out situations in which the delay is 0, meaning the incident that
+# occurred didn't affect service
[…]

### 2025-06-19 scripts/03-parse_codes.R

@@ -1,190 +1,0 @@
-#### Preamble ####
-# Purpose: Group delay data into similar sets of incidents
-# Author: Timothius Prajogi
-# Date: 23 January 2024
-# Contact: tim.prajogi@mail.utoronto.ca
-# License: MIT
-# Prerequisites: 01-download_data, 02-clean_data
-
-#### Workspace setup ####
-library(tidyverse)
-library(janitor)
-
-### Bus data ###
-
-# Load in non-grouped data
-bus_delay_data <- 
-  read_csv(
-    file = "inputs/data/bus_delay_statistics.csv",
-    show_col_types = FALSE
-  )
-
-# Group bus incidents
-cleaned_bus_delay_data <- 
-  bus_delay_data |>
-  mutate(
-    incident =
-      case_match(
-        incident,
-        "Diversion" ~ "Miscellaneous",
-        "Security" ~ "Security/Safety",
-        "Cleaning - Unsanitary" ~ "Security/Safety",
-        "Emergency Services" ~ "Security/Safety",
-        "Collision - TTC" ~ "Operator",
-        "Mechanical" ~ "Equipment/Mechanical",
-        "Operations - Operator" ~ "Operator",
-        "Investigation" ~ "Security/Safety",
-        "Utilized Off Route" ~ "Miscellaneous",
-        "General Delay" ~ "Miscellaneous",
-        "Road Blocked - NON-TTC Collision" ~ "Miscellaneous",
-        "Held By" ~ "Miscellaneous",
-        "Vision" ~ "Security/Safety"
-      )
-  )
-
-# Save cleaned data
-write_csv(cleaned_bus_delay_data, 'outputs/data/cleaned_bus_delay_statistics.csv')
-
-### SUBWAY DATA ###
-
-# Read in subway codes
-subway_codes <-
-  read_csv(
-    file = "inputs/data/subway_delay_codes.csv",
-    show_col_types = FALSE
-  )
-
-
-# Section codes into incident groups
-cleaned_subway_codes <- 
-  subway_codes |>
-  mutate(
-    incident = case_when(
-      startsWith(code, "E") ~ "Equipment/Mechanical",
-      startsWith(code, "MUI") ~ "Security/Safety",
-      startsWith(code, "MUS") ~ "Security/Safety",
-      startsWith(code, "MUP") ~ "Security/Safety",
-      startsWith(code, "MUD") ~ "Equipment/Mechanical",
[…]

### 2025-06-19 scripts/03-parse_codes.py

@@ -1,0 +1,143 @@
+#### Preamble ####
+# Purpose: Group delay data into similar sets of incidents
+# Author: Timothius Prajogi
+# Date: 23 January 2024
+# Contact: tim.prajogi@mail.utoronto.ca
+# License: MIT
+# Prerequisites: 01-download_data, 02-clean_data
+
+#### Workspace setup ####
+import pandas as pd
+
+### Bus data ###
+
+# Load in non-grouped data
+bus_delay_data = pd.read_csv("inputs/data/bus_delay_statistics.csv")
+
+# Group bus incidents
+bus_incident_map = {
+    "Diversion": "Miscellaneous",
+    "Security": "Security/Safety",
+    "Cleaning - Unsanitary": "Security/Safety",
+    "Emergency Services": "Security/Safety",
+    "Collision - TTC": "Operator",
+    "Mechanical": "Equipment/Mechanical",
+    "Operations - Operator": "Operator",
+    "Investigation": "Security/Safety",
+    "Utilized Off Route": "Miscellaneous",
+    "General Delay": "Miscellaneous",
+    "Road Blocked - NON-TTC Collision": "Miscellaneous",
+    "Held By": "Miscellaneous",
+    "Vision": "Security/Safety"
+}
+
+cleaned_bus_delay_data = bus_delay_data.copy()
+cleaned_bus_delay_data["incident"] = cleaned_bus_delay_data["incident"].map(bus_incident_map)
+
+# Save cleaned data
+cleaned_bus_delay_data.to_csv("outputs/data/cleaned_bus_delay_statistics.csv", index=False)
+
+### SUBWAY DATA ###
+
+# Read in subway codes
+subway_codes = pd.read_csv("inputs/data/subway_delay_codes.csv")
+
+# Section codes into incident groups
+def classify_subway_incident(code):
+    if str(code).startswith("E"):
+        return "Equipment/Mechanical"
+    if str(code).startswith(("MUI", "MUS", "MUP")):
+        return "Security/Safety"
+    if str(code).startswith("MUD") or str(code).startswith("MUE"):
+        return "Equipment/Mechanical"
+    if str(code).startswith("P"):
+        return "Equipment/Mechanical"
+    if str(code).startswith("S"):
+        return "Security/Safety"
+    if str(code).startswith("T"):
+        return "Operator"
+    return "Miscellaneous"
+
[…]

### 2025-06-19 scripts/04-summarise_data.R

@@ -1,111 +1,0 @@
-#### Preamble ####
-# Purpose: Group delay data into similar sets of incidents
-# Author: Timothius Prajogi
-# Date: 23 January 2024
-# Contact: tim.prajogi@mail.utoronto.ca
-# License: MIT
-# Prerequisites: 01-download_data, 02-clean_data, 03-parse_codes
-
-#### Workspace setup ####
-library(tidyverse)
-
-# Read in bus data
-bus_delay_data <- 
-  read_csv(
-    file = "outputs/data/cleaned_bus_delay_statistics.csv",
-    show_col_types = FALSE
-  )
-
-# Enforce ordering of days
-bus_delay_data$day <- factor(bus_delay_data$day, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
-
-# Summarize into average number of delays per day of the week (Mon, Tues, ...)
-avg_num_bus_delays_by_day <-
-  bus_delay_data |> 
-  group_by(date, day) |> 
-  summarise(n = n(), .groups = "keep") |>
-  group_by(day) |>
-  summarise(mean_num_delays = mean(n))
-
-# Summarize into total delay time and number of delays per date (Jan 1st, 2nd, ...)
-total_bus_delay_time_by_date <-
-  bus_delay_data |> 
-  group_by(date, day) |> 
-  summarise(
-    total_delay_time = sum(min_delay), 
-    n = n(), 
-    mean_delay_time = (total_delay_time / n),
-    .groups = "keep")
-
-# Summarize into the total number of delays per incident type (e.g. Equipment/Mechanical)
-total_num_bus_delays_by_incident <-
-  bus_delay_data |>
-  group_by(incident) |>
-  summarise(n = n())
-
-# Read in subway data
-subway_delay_data <- 
-  read_csv(
-    file = "outputs/data/cleaned_subway_delay_statistics.csv",
-    show_col_types = FALSE
-  )
-
-# Enforce ordering of days
-subway_delay_data$day <- factor(subway_delay_data$day, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
-
-# Summarize into average number of delays per day of the week (Mon, Tues, ...)
-avg_num_subway_delays_by_day <-
-  subway_delay_data |> 
-  group_by(date, day) |> 
-  summarise(n = n(), .groups = "keep") |>
-  group_by(day) |>
[…]

### 2025-06-19 scripts/04-summarise_data.py

@@ -1,0 +1,114 @@
+#### Preamble ####
+# Purpose: Group delay data into similar sets of incidents
+# Author: Timothius Prajogi
+# Date: 23 January 2024
+# Contact: tim.prajogi@mail.utoronto.ca
+# License: MIT
+# Prerequisites: 01-download_data, 02-clean_data, 03-parse_codes
+
+#### Workspace setup ####
+
+import pandas as pd
+
+# Ordered day list for consistent grouping
+ordered_days = ["Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"]
+
+# Read in bus data
+bus_delay_data = pd.read_csv("outputs/data/cleaned_bus_delay_statistics.csv")
+bus_delay_data["day"] = pd.Categorical(bus_delay_data["day"], categories=ordered_days, ordered=True)
+
+# Summarize into average number of delays per day of the week (Mon, Tues, ...)
+avg_num_bus_delays_by_day = (
+    bus_delay_data.groupby(["date", "day"], observed=False)
+    .size()
+    .reset_index(name="n")
+    .groupby("day", observed=False)
+    .agg(mean_num_delays=("n", "mean"))
+    .reset_index()
+)
+
+# Summarize into total delay time and number of delays per date (Jan 1st, 2nd, ...)
+total_bus_delay_time_by_date = (
+    bus_delay_data.groupby(["date", "day"], observed=False)
+    .agg(
+        total_delay_time=("min_delay", "sum"),
+        n=("min_delay", "size")
+    )
+    .reset_index()
+)
+total_bus_delay_time_by_date["mean_delay_time"] = (
+    total_bus_delay_time_by_date["total_delay_time"] / total_bus_delay_time_by_date["n"]
+)
+
+# Summarize into the total number of delays per incident type
+total_num_bus_delays_by_incident = (
+    bus_delay_data.groupby("incident", observed=False)
+    .size()
+    .reset_index(name="n")
+)
+
+# Read in subway data
+subway_delay_data = pd.read_csv("outputs/data/cleaned_subway_delay_statistics.csv")
+subway_delay_data["day"] = pd.Categorical(subway_delay_data["day"], categories=ordered_days, ordered=True)
+
+# Summarize into average number of delays per day of the week (Mon, Tues, ...)
+avg_num_subway_delays_by_day = (
+    subway_delay_data.groupby(["date", "day"], observed=False)
+    .size()
+    .reset_index(name="n")
[…]

### 2025-06-20 outputs/paper.qmd
fixed issue -- papr.qmd in python now

@@
 
-```{r}
+```{python}
 #| echo: false
@@
 #### Workspace setup ####
-library(janitor)
-library(knitr)
-library(opendatatoronto)
-library(tidyverse)
+import matplotlib.pyplot as plt
+import pandas as pd
 
@@
 
-# Read in cleaned data
-bus_delay_data <- 
-  read_csv(
-     file = "data/cleaned_bus_delay_statistics.csv",
-     show_col_types = FALSE
-  )
+# Read bus delay data
+bus_delay_data = pd.read_csv("data/cleaned_bus_delay_statistics.csv")
 
-subway_delay_data <-
-  read_csv(
-     file = "data/cleaned_subway_delay_statistics.csv",
-     show_col_types = FALSE
-  )
+# Read subway delay data
+subway_delay_data = pd.read_csv("data/cleaned_subway_delay_statistics.csv")
 ```
@@
 # Display a sample of the bus dataset in a table
-bus_delay_data |>
-  head(5) |>
-  kable(
-    col.names = c("Date", "Time", "Day", "Incident", "Min Delay", "Min Gap"),
-    booktabs = TRUE
-  )
+sample_bus_delay_data = bus_delay_data.head(5)
+sample_bus_delay_data.columns = ["Date", "Time", "Day", "Incident", "Min Delay", "Min Gap"]
+sample_bus_delay_data
 ```
 
-```{r}
+```{python}
 #| echo: false
@@
 # Display a sample of the subway dataset in a table
-subway_delay_data |>
-  head(5) |>
-  kable(
-    col.names = c("Date", "Time", "Day", "Incident", "Min Delay", "Min Gap", "Line"),
-    booktabs = TRUE
-  )
+sample_subway_delay_data = subway_delay_data.head(5)
+sample_subway_delay_data.columns = ["Date", "Time", "Day", "Incident", "Min Delay", "Min Gap", "Line"]
+sample_subway_delay_data
 
@@
 
-```{r}
+```{python}
 #| eval: true
@@
 
-# Display the total amount of delays (on any date) for
-# a particular time
+# Convert time column to datetime if not already
+bus_delay_data["time"] = pd.to_datetime(bus_delay_data["time"], errors="coerce")
 
-bus_delay_data |>
-  ggplot(aes(x = time)) +
-  geom_histogram(fill = "blue", alpha = 0.75) +
-  scale_x_time(breaks = scales::pretty_breaks(n = 12)) +
-  xlab("Time") +
-  ylab("Count") +
-  theme_light() +
-  theme(axis.text.x = element_text(angle = 60, hjust = 1))
+# Drop any rows with invalid times
+valid_time_data = bus_delay_data.dropna(subset=["time"])
+
+# Extract just the time part for histogram binning
+valid_time_data["hour"] = valid_time_data["time"].dt.hour + valid_time_data["time"].dt.minute / 60
+
+# Plot
+plt.figure(figsize=(8, 4))
+plt.hist(valid_time_data["hour"], bins=24, color="blue", alpha=0.75)
+
+plt.xlabel("Time")
+plt.ylabel("Count")
+plt.xticks(ticks=range(0, 25, 2), labels=[f"{h}:00" for h in range(0, 25, 2)], rotation=60, ha="right")
+plt.grid(axis="y", linestyle=":", linewidth=0.5)
+plt.tight_layout()
+plt.show()
 
@@
 
-# Display the total amount of delays (on any date) for
-# a particular time
+# Ensure the "time" column is interpreted as a datetime
+subway_delay_data["time"] = pd.to_datetime(subway_delay_data["time"], errors="coerce")
 
-subway_delay_data |>
-  ggplot(aes(x=time)) +
-  geom_histogram(fill = "red", alpha = 0.75) +
-  scale_x_time(breaks = scales::pretty_breaks(n = 12)) +
-  xlab("Time") +
-  ylab("Count") +
-  theme_light() + 
-  theme(axis.text.x = element_text(angle = 60, hjust = 1))
+# Remove rows with invalid times
+valid_times = subway_delay_data.dropna(subset=["time"])
+
+# Convert times to fractional hours for histogram binning
+valid_times["hour_decimal"] = (
+    valid_times["time"].dt.hour + valid_times["time"].dt.minute / 60
+)
+
+# Plot histogram
+plt.figure(figsize=(8, 4))
+plt.hist(valid_times["hour_decimal"], bins=24, color="red", alpha=0.75)
+
+plt.xlabel("Time")
+plt.ylabel("Count")
+plt.xticks(
+    ticks=range(0, 25, 2),
+    labels=[f"{h}:00" for h in range(0, 25, 2)],
+    rotation=60,
+    ha="right"
+)
+plt.grid(axis="y", linestyle=":", linewidth=0.5)
+plt.tight_layout()
+plt.show()
 
@@
 
-# Load in summarized data
-avg_num_bus_delays_by_day <- read_csv(
-   file = "data/summaries/avg_num_bus_delays_by_day.csv",
-   show_col_types = FALSE
+# Define day order
+ordered_days = ["Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"]
+
+# Load and format bus data
+avg_num_bus_delays_by_day = pd.read_csv("data/summaries/avg_num_bus_delays_by_day.csv")
+avg_num_bus_delays_by_day["day"] = pd.Categorical(avg_num_bus_delays_by_day["day"], categories=ordered_days, ordered=True)
+avg_num_bus_delays_by_day = avg_num_bus_delays_by_day.sort_values("day")
+
+# Plot bus delays
+plt.figure(figsize=(8, 4))
+plt.bar(
+    avg_num_bus_delays_by_day["day"],
+    avg_num_bus_delays_by_day["mean_num_delays"],
+    color="skyblue"
 )
+plt.xlabel("Day of the Week")
+plt.ylabel("Mean Number of Delays")
+plt.xticks(rotation=60, ha="right")
+plt.grid(axis="y", linestyle=":", linewidth=0.5)
+plt.tight_layout()
+plt.show()
 
-avg_num_bus_delays_by_day$day <- 
-  factor(avg_num_bus_delays_by_day$day, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
+# Load and format subway data
+avg_num_subway_delays_by_day = pd.read_csv("data/summaries/avg_num_subway_delays_by_day.csv")
+avg_num_subway_delays_by_day["day"] = pd.Categorical(avg_num_subway_delays_by_day["day"], categories=ordered_days, ordered=True)
+avg_num_subway_delays_by_day = avg_num_subway_delays_by_day.sort_values("day")
 
-avg_num_bus_delays_by_day |>
-  ggplot(aes(x = day, y = mean_num_delays, fill = day)) +
-  geom_bar(stat = 'identity') +
-  xlab("Day of the Week") +
-  ylab("Mean Number of Delays") + 
-  theme_light() +
-  theme(axis.text.x = element_text(angle = 60, hjust = 1))
-
-avg_num_subway_delays_by_day <- read_csv(
-   file = "data/summaries/avg_num_subway_delays_by_day.csv",
-   show_col_types = FALSE
+# Plot subway delays
+plt.figure(figsize=(8, 4))
+plt.bar(
+    avg_num_subway_delays_by_day["day"],
+    avg_num_subway_delays_by_day["mean_num_delays"],
+    color="salmon"
 )
-
-avg_num_subway_delays_by_day$day <- 
-  factor(avg_num_subway_delays_by_day$day, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
-
-avg_num_subway_delays_by_day |>
-  ggplot(aes(x = day, y = mean_num_delays, fill = day)) +
-  geom_bar(stat = 'identity') +
-  xlab("Day of the Week") +
-  ylab("Mean Number of Delays") + 
-  theme_light() +
-  theme(axis.text.x = element_text(angle = 60, hjust = 1))
+plt.xlabel("Day of the Week")
+plt.ylabel("Mean Number of Delays")
+plt.xticks(rotation=60, ha="right")
+plt.grid(axis="y", linestyle=":", linewidth=0.5)
+plt.tight_layout()
+plt.show()
 
@@
 
-total_bus_delay_time_by_date <- 
-  read_csv(
-   file = "data/summaries/total_bus_delay_time_by_date.csv",
-   show_col_types = FALSE
-)
+# Load summarized bus delay data
+total_bus_delay_time_by_date = pd.read_csv("data/summaries/total_bus_delay_time_by_date.csv")
 
-# Take 5 random rows from the summarized data 
-set.seed(302)
-total_bus_delay_time_by_date[sample(nrow(total_bus_delay_time_by_date), 5), ] |>
-  kable(
-    col.names = c("Date", "Day", "Total Delay Time", "Delay Count", "Average Delay Time"),
-    booktabs = TRUE
-  )
+# Take 5 random rows with a fixed seed
+sample_bus_delays = total_bus_delay_time_by_date.sample(n=5, random_state=302)
+
+# Rename columns to match desired display names
+sample_bus_delays.columns = [
+    "Date",
+    "Day",
+    "Total Delay Time",
+    "Delay Count",
+    "Average Delay Time"
+]
+
+# Display the sample table
+sample_bus_delays
 
@@
 
-total_bus_delay_time_by_date |>
-  ggplot(aes(x = date, y = total_delay_time)) + 
-  geom_point() +
-  geom_smooth(method = "loess") +
-  theme_light() +
-  xlab("Date") +
-  ylab("Total Delay Time (Minutes)")
+plt.figure(figsize=(10, 4))
+plt.scatter(
+    total_bus_delay_time_by_date["date"],
+    total_bus_delay_time_by_date["total_delay_time"],
+    color="blue", alpha=0.6
+)
+plt.xlabel("Date")
+plt.ylabel("Total Delay Time (Minutes)")
+plt.grid(axis="y", linestyle=":", linewidth=0.5)
+plt.tight_layout()
+plt.show()
 
-total_bus_delay_time_by_date |>
-  ggplot(aes(x = date, y = mean_delay_time)) + 
-  geom_point() +
-  geom_smooth(method = "loess") +
-  theme_light() +
-  xlab("Date") +
-  ylab("Total Delay Time (Minutes)")
+plt.figure(figsize=(10, 4))
+plt.scatter(
+    total_bus_delay_time_by_date["date"],
+    total_bus_delay_time_by_date["mean_delay_time"],
+    color="green", alpha=0.6
+)
+plt.xlabel("Date")
+plt.ylabel("Total Delay Time (Minutes)")
+plt.grid(axis="y", linestyle=":", linewidth=0.5)
+plt.tight_layout()
+plt.show()
 
@@
 
-# Load summarized data about number of delays vs incidents
+# Load summarized data
+bus_delays_by_incident = pd.read_csv("data/summaries/total_num_bus_delays_by_incident.csv")
 
-bus_delays_by_incident <- read_csv(
-   file = "data/summaries/total_num_bus_delays_by_incident.csv",
-   show_col_types = FALSE
+# Plot
+plt.figure(figsize=(8, 4))
+bars = plt.bar(
+    bus_delays_by_incident["incident"],
+    bus_delays_by_incident["n"],
+    color=plt.cm.Set2.colors  # distinct fill colors
 )
-bus_delays_by_incident |> 
-  ggplot(aes(x=incident, y=n, fill=incident)) +
-  geom_bar(stat = "identity") +
-  xlab("Incident Type") +
-  ylab("Delay Count") +
-  theme_light() +
-  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
-  guides(fill=guide_legend(title="Incident Type"))
+
+plt.xlabel("Incident Type")
+plt.ylabel("Delay Count")
+plt.xticks(rotation=60, ha="right")
+plt.grid(axis="y", linestyle=":", linewidth=0.5)
+
+# Add legend manually
+incident_labels = bus_delays_by_incident["incident"].tolist()
+plt.legend(bars, incident_labels, title="Incident Type", bbox_to_anchor=(1.05, 1), loc="upper left")
+
+plt.tight_layout()
+plt.show()
 
@@
 
-subway_delays_by_incident <- read_csv(
-   file = "data/summaries/total_num_subway_delays_by_incident.csv",
-   show_col_types = FALSE
+# Load summarized data
+subway_delays_by_incident = pd.read_csv("data/summaries/total_num_subway_delays_by_incident.csv")
+
+# Plot
+plt.figure(figsize=(8, 4))
+bars = plt.bar(
+    subway_delays_by_incident["incident"],
+    subway_delays_by_incident["n"],
+    color=plt.cm.Set2.colors
 )
-subway_delays_by_incident |> 
-  ggplot(aes(x = incident, y = n, fill = incident)) +
-  geom_bar(stat = "identity") +
-  xlab("Incident Type") +
-  ylab("Delay Count") +
-  theme_light() +
-  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
-  guides(fill=guide_legend(title="Incident Type"))
+
+plt.xlabel("Incident Type")
+plt.ylabel("Delay Count")
+plt.xticks(rotation=60, ha="right")
+plt.grid(axis="y", linestyle=":", linewidth=0.5)
+
+# Add legend
+incident_labels = subway_delays_by_incident["incident"].tolist()
+plt.legend(bars, incident_labels, title="Incident Type", bbox_to_anchor=(1.05, 1), loc="upper left")
+
+plt.tight_layout()
+plt.show()
 
@@
 
-subway_delays_by_line <- read_csv(
-   file = "data/summaries/total_num_subway_delays_by_line.csv",
-   show_col_types = FALSE
+# Load summarized data
+subway_delays_by_line = pd.read_csv("data/summaries/total_num_subway_delays_by_line.csv")
+
+# Plot
+plt.figure(figsize=(8, 4))
+bars = plt.bar(
+    subway_delays_by_line["line"],
+    subway_delays_by_line["n"],
+    color=plt.cm.Set2.colors  # visually distinct colors
 )
 
-subway_delays_by_line |> 
-  ggplot(aes(x = line, y = n, fill = line)) +
-  geom_bar(stat = "identity") + 
-  theme_light() +
-  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
-  xlab("TTC Line") +
-  ylab("Number of Delays")
+plt.xlabel("TTC Line")
+plt.ylabel("Number of Delays")
+plt.xticks(rotation=60, ha="right")
+plt.grid(axis="y", linestyle=":", linewidth=0.5)
+
+# Optional: Add legend
+line_labels = subway_delays_by_line["line"].tolist()
+plt.legend(bars, line_labels, title="TTC Line", bbox_to_anchor=(1.05, 1), loc="upper left")
+
+plt.tight_layout()
+plt.show()
 

### 2025-06-20 scripts/04-summarise_data.py

@@ -7,88 +7,89 @@
 # Prerequisites: 01-download_data, 02-clean_data, 03-parse_codes
 
 #### Workspace setup ####
-
 import pandas as pd
-
-# Ordered day list for consistent grouping
-ordered_days = ["Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"]
 
 # Read in bus data
 bus_delay_data = pd.read_csv("outputs/data/cleaned_bus_delay_statistics.csv")
-bus_delay_data["day"] = pd.Categorical(bus_delay_data["day"], categories=ordered_days, ordered=True)
 
-# Summarize into average number of delays per day of the week (Mon, Tues, ...)
+# Ensure date is just the calendar date (not datetime with time)
+bus_delay_data["date"] = pd.to_datetime(bus_delay_data["date"]).dt.date
+
+
 avg_num_bus_delays_by_day = (
-    bus_delay_data.groupby(["date", "day"], observed=False)
+    bus_delay_data
+    .groupby(["date", "day"], observed=False)  # observed=False ensures all day levels are preserved
     .size()
-    .reset_index(name="n")
-    .groupby("day", observed=False)
-    .agg(mean_num_delays=("n", "mean"))
-    .reset_index()
+    .groupby(level="day", observed=False)
+    .mean()
+    .reset_index(name="mean_num_delays")
 )
 
 # Summarize into total delay time and number of delays per date (Jan 1st, 2nd, ...)
 total_bus_delay_time_by_date = (
-    bus_delay_data.groupby(["date", "day"], observed=False)
-    .agg(
-        total_delay_time=("min_delay", "sum"),
-        n=("min_delay", "size")
-    )
+    bus_delay_data
+    .groupby(["date", "day"])
+    .agg(total_delay_time=("min_delay", "sum"), n=("min_delay", "count"))
     .reset_index()
 )
-total_bus_delay_time_by_date["mean_delay_time"] = (
-    total_bus_delay_time_by_date["total_delay_time"] / total_bus_delay_time_by_date["n"]
-)
+total_bus_delay_time_by_date["mean_delay_time"] = total_bus_delay_time_by_date["total_delay_time"] / total_bus_delay_time_by_date["n"]
 
-# Summarize into the total number of delays per incident type
+# Summarize into the total number of delays per incident type (e.g. Equipment/Mechanical)
 total_num_bus_delays_by_incident = (
-    bus_delay_data.groupby("incident", observed=False)
[…]

### 2025-06-24 outputs/paper.qmd
anonymized, changed citations, removed files, changed title

@@
 ---
-title: "Analyzing TTC Delay Data to Improve Future Efficiency"
+title: "Using TTC Delay Data to Identify Opportunities for Improved Efficiency"
 abstract: "TTC delays have been increasing over time, and especially now in a period of overhaul in many parts of the system, it becomes important to analyze the effects of subway and bus delays in relation with each other as well as commuter loads.
@@
 Date and time factors are out of anyones control, but must be accounted for to improve future operational efficieny, while equipment quality has also been shown to be important, with mechanical failures heavily decreasing the reliability of buses."
-thanks: "Code and data supporting this analysis is available at: https://github.com/prajogt/ttc_analysis.git"
-author: "Timothius Prajogi"
+thanks: "Code and data supporting this analysis is available at: https://github.com/EveHughes/ttc_analysis.git"
+author: "Omar Hartono"
 date: today
@@
 
-These delay statistics were downloaded, cleaned, parsed, analyzed, and visualized using R [@citeR], a statistical programming language, with package support from `tidyverse` [@citetidyverse], a collection of libraries which included the following packages that were utilized:
+These delay statistics were downloaded, cleaned, parsed, analyzed, and visualized using Python [@python], and a collection of libraries including the following packages that were utilized:
 
-* `ggplot2` [@citeggplot] 
-* `dplyr` [@citedplyr] 
-* `readr` [@citereadr] 
-* `tibble` [@citetibble]
-
-For additional assistance with cleaning, the `janitor` [@citejanitor] package was used. For additional assistance with report generation the `knitr` [@citeknitr] package was used.
+* `numpy` [@numpy] 
+* `matplotlib` [@matplotlib] 
+* `pandas` [@pandas] 
+.
 

### 2025-06-24 scripts/00-simulate_data.py

@@ -1,8 +1,8 @@
 #### Preamble ####
 # Purpose: Simulate delay data
-# Author: Timothius Prajogi
+# Author: Daniel Hartono
 # Date: 23 January 2024
-# Contact: tim.prajogi@mail.utoronto.ca
+# Contact: daniel.hartono@mail.utoronto.ca
 # License: MIT
 # Prerequisites: None
 

### 2025-06-24 scripts/01-download_data.py

@@ -1,8 +1,8 @@
 #### Preamble ####
 # Purpose: Download TTC delay data from opendatatoronto
-# Author: Timothius Prajogi
+# Author: Daniel Hartono
 # Date: 23 January 2024
-# Contact: tim.prajogi@mail.utoronto.ca
+# Contact: daniel.hartono@mail.utoronto.ca
 # License: MIT
 # Prerequisites: none
 

### 2025-06-24 scripts/02-clean_data.py

@@ -1,8 +1,8 @@
 #### Preamble ####
 # Purpose: Clean raw datasets
-# Author: Timothius Prajogi
+# Author: Daniel Hartono
 # Date: 23 January 2024
-# Contact: tim.prajogi@mail.utoronto.ca
+# Contact: daniel.hartono@mail.utoronto.ca
 # License: MIT
 # Prerequisites: 01-download_data
 

### 2025-06-24 scripts/03-parse_codes.py

@@ -1,8 +1,8 @@
 #### Preamble ####
 # Purpose: Group delay data into similar sets of incidents
-# Author: Timothius Prajogi
+# Author: Daniel Hartono
 # Date: 23 January 2024
-# Contact: tim.prajogi@mail.utoronto.ca
+# Contact: daniel.hartono@mail.utoronto.ca
 # License: MIT
 # Prerequisites: 01-download_data, 02-clean_data
 

### 2025-06-24 scripts/04-summarise_data.py

@@ -1,8 +1,8 @@
 #### Preamble ####
 # Purpose: Group delay data into similar sets of incidents
-# Author: Timothius Prajogi
+# Author: Daniel Hartono
 # Date: 23 January 2024
-# Contact: tim.prajogi@mail.utoronto.ca
+# Contact: daniel.hartono@mail.utoronto.ca
 # License: MIT
 # Prerequisites: 01-download_data, 02-clean_data, 03-parse_codes
 
